---
title: "Adverse Events"
order: 2
---


```{r setup script, include=FALSE, purl=FALSE}
invisible_hook_purl <- function(before, options, ...) {
  knitr::hook_purl(before, options, ...)
  NULL
}
knitr::knit_hooks$set(purl = invisible_hook_purl)
```

## Introduction

This guide will show you how pharmaverse packages, along with some from tidyverse, can be used to create an Adverse Events table, using the `{pharmaverseadam}` `ADSL` and `ADAE` data as an input.

In the examples below, we illustrate two general approaches for creating an Adverse events table.  The first is the classic method of creating summary tables directly from a data set. The second utilizes Analysis Results Datasets—part of the emerging CDISC Analysis Results Standard.

## {rtables} & {tern}

The packages used with a brief description of their purpose are as follows:

* [`{rtables}`](https://insightsengineering.github.io/rtables/): designed to create and display complex tables with R.
* [`{tern}`](https://insightsengineering.github.io/tern/): contains analysis functions to create tables and graphs used for clinical trial reporting.

### Load Data and Required pharmaverse Package

After installation of packages, the first step is to load our pharmaverse packages and input data. Here, we are going to encode missing entries in a data frame `adsl` and `adae`.

Note that `{tern}` depends on `{rtables}` so the latter is automatically attached.

```{r setup, message=FALSE, warning=FALSE, results='hold'}
library(pharmaverseadam)
library(tern)
library(dplyr)

adsl <- adsl %>%
  df_explicit_na()

adae <- adae %>%
  df_explicit_na()
```

### Start preprocessing

Now we will add some pre-processing to add labels ready for display in the table and how the output will be split.

```{r preproc}
adae <- adae %>%
  var_relabel(
    AEBODSYS = "MedDRA System Organ Class",
    AEDECOD = "MedDRA Preferred Term"
  ) %>%
  filter(SAFFL == "Y")

# Define the split function
split_fun <- drop_split_levels
```

### Adverse Events table

Now we create the Adverse Events table.

```{r table}
lyt <- basic_table(show_colcounts = TRUE) %>%
  split_cols_by(var = "ACTARM") %>%
  add_overall_col(label = "All Patients") %>%
  analyze_num_patients(
    vars = "USUBJID",
    .stats = c("unique", "nonunique"),
    .labels = c(
      unique = "Total number of patients with at least one adverse event",
      nonunique = "Overall total number of events"
    )
  ) %>%
  split_rows_by(
    "AEBODSYS",
    child_labels = "visible",
    nested = FALSE,
    split_fun = split_fun,
    label_pos = "topleft",
    split_label = obj_label(adae$AEBODSYS)
  ) %>%
  summarize_num_patients(
    var = "USUBJID",
    .stats = c("unique", "nonunique"),
    .labels = c(
      unique = "Total number of patients with at least one adverse event",
      nonunique = "Total number of events"
    )
  ) %>%
  count_occurrences(
    vars = "AEDECOD",
    .indent_mods = -1L
  ) %>%
  append_varlabels(adae, "AEDECOD", indent = 1L)

result <- build_table(lyt, df = adae, alt_counts_df = adsl)

result
```

## {tfrmt} & {cards}

In the example below, we will use the [{tfrmt}](https://github.com/GSK-Biostatistics/tfrmt) and [{cards}](https://insightsengineering.github.io/cards/) packages to create a adverse events tables.

- The {cards} package creates Analysis Results Datasets (ARDs, which are a part of the [CDISC Analysis Results Standard](https://www.cdisc.org/standards/foundational/analysis-results-standard)).
- The {tfrmt} utilizes ARDs to create tables.

In the example below, we first build an ARD with the needed summary statistics using {cards}.
Then, we use the ARD to build the adverse events table with {tfrmt}.

### Load libraries

```{r message=FALSE, warning=FALSE}
library(pharmaverseadam) # for clinical trial data
library(dplyr) # for data manipulation
library(cards) # for creating analysis result displays
library(tfrmt) # for formatting tables in R
```

------------------------------------------------------------------------

### Step 1: Import data

Subset to safety population. 

```{r}
# Filter to include only subjects marked as part of the safety population
adsl <- pharmaverseadam::adsl |>
  filter(SAFFL == "Y")
# Load adverse event data
adae <- pharmaverseadam::adae |>
  filter(SAFFL == "Y" & TRTEMFL == "Y")
```

------------------------------------------------------------------------

### Step 2: Data Preparation

#### Analysis

Create an Analysis Result Display (ARD) using the {cards} package.

```{r}
# Create an ARD that stacks hierarchical data of adverse events
# Grouping by treatment, system organ class, and preferred term
ae_ard <- ard_stack_hierarchical(
  data = adae,
  by = TRT01A, # Note: by variables must be present in the denominator dataset
  variables = c(AEBODSYS, AETERM),
  statistic = ~ c("n", "p"), # Calculate count and percentage
  denominator = adsl,
  id = USUBJID,
  over_variables = TRUE,
  overall = TRUE
)

# Filter adae and adsl with trt01a set to "Total" and create a new ARD for the total column
adae2 <- adae |>
  mutate(TRT01A = "Total")
adsl2 <- adsl |>
  mutate(TRT01A = "Total")

ae2_ard <- ard_stack_hierarchical(
  data = adae2,
  by = TRT01A, # Note: by variables must be present in the denominator dataset
  variables = c(AEBODSYS, AETERM),
  denominator = adsl2,
  statistic = ~ c("n", "p"),
  id = USUBJID,
  over_variables = TRUE,
  overall = TRUE
) |>
  filter(group2 == "TRT01A" | variable == "TRT01A") # filter to stats we need
```


### Tidy for table


Further refine the ARD for use in tfrmt

```{r}
ae3_ard <- bind_ard(ae_ard, ae2_ard) |>
  # reshape the data
  shuffle_card(fill_hierarchical_overall = "ANY EVENT") |>
  # transform group-level freqs/pcts into a singular "bigN" row
  prep_big_n(vars = "TRT01A") |>
  # for nested variables, fill any missing values with "ANY EVENT"
  prep_hierarchical_fill(vars = c("AEBODSYS", "AETERM"), fill = "ANY EVENT") |>
  mutate(TRT01A = ifelse(TRT01A == "Overall TRT01A", "Total", TRT01A))

# create ordering columns, sort by AEBODSYS
ordering_aebodsys <- ae3_ard |>
  filter(TRT01A == "Total", stat_name == "n", AETERM == "ANY EVENT") |>
  arrange(desc(stat)) |>
  mutate(ord1 = row_number()) |>
  select(AEBODSYS, ord1)

# sort by AETERM after AEBODSYS order
ordering_aeterm <- ae3_ard |>
  filter(TRT01A == "Total", stat_name == "n") |>
  group_by(AEBODSYS) |>
  arrange(desc(stat)) |>
  mutate(ord2 = row_number()) |>
  select(AEBODSYS, AETERM, ord2)

# join on our ordering columns and keep required columns
ae4_ard <- ae3_ard |>
  full_join(ordering_aebodsys, by = "AEBODSYS") |>
  full_join(ordering_aeterm, by = c("AEBODSYS", "AETERM")) |>
  select(AEBODSYS, AETERM, ord1, ord2, stat, stat_name, TRT01A)
```

------------------------------------------------------------------------

### Step 3: Creating the `{tfrmt}` Table

```{r}
AE_T01 <- tfrmt_n_pct(
  n = "n", pct = "p",
  pct_frmt_when = frmt_when(
    "==1" ~ frmt("(100%)"),
    ">=0.995" ~ frmt("(>99%)"),
    "==0" ~ frmt(""),
    "<=0.01" ~ frmt("(<1%)"),
    "TRUE" ~ frmt("(xx.x%)", transform = ~ . * 100)
  )
) |>
  tfrmt(
    group = AEBODSYS,
    label = AETERM,
    param = stat_name,
    value = stat,
    column = TRT01A,
    sorting_cols = c(ord1, ord2),
    col_plan = col_plan(
      "System Organ Class
            Preferred Term" = AEBODSYS, Placebo, `Xanomeline High Dose`, `Xanomeline Low Dose`,
      -ord1, -ord2
    ),
    row_grp_plan = row_grp_plan(row_grp_structure(
      group_val = ".default", element_block(post_space = " ")
    )),
    big_n = big_n_structure(param_val = "bigN", n_frmt = frmt(" (N=xx)"))
  ) |>
  print_to_gt(ae4_ard)

AE_T01
```
