---
title: "Adverse Events"
order: 2
---

## Introduction

This guide will show you how pharmaverse packages, along with some from tidyverse, can be used to create an Adverse Events table, using the `{pharmaverseadam}` `ADSL` and `ADAE` data as an input.

The packages used with a brief description of their purpose are as follows:

* [`{rtables}`](https://insightsengineering.github.io/rtables/): designed to create and display complex tables with R.
* [`{tern}`](https://insightsengineering.github.io/tern/): contains analysis functions to create tables and graphs used for clinical trial reporting.

## Load Data and Required pharmaverse Package

After installation of packages, the first step is to load our pharmaverse packages and input data. Here, we are going to encode missing entries in a data frame `adsl` and `adae`.

Note that `{tern}` depends on `{rtables}` so the latter is automatically attached.

```{r setup, message=FALSE, warning=FALSE, results='hold'}
library(pharmaverseadam)
library(tern)
library(dplyr)

adsl <- adsl %>%
  df_explicit_na()

adae <- adae %>%
  df_explicit_na()
```

## Start preprocessing

Now we will add some pre-processing to add labels ready for display in the table and how the output will be split.

```{r preproc}
adae <- adae %>%
  var_relabel(
    AEBODSYS = "MedDRA System Organ Class",
    AEDECOD = "MedDRA Preferred Term"
  ) %>%
  filter(SAFFL == "Y")

# Define the split function
split_fun <- drop_split_levels
```

## Adverse Events table

Now we create the Adverse Events table.

```{r table}
lyt <- basic_table(show_colcounts = TRUE) %>%
  split_cols_by(var = "ACTARM") %>%
  add_overall_col(label = "All Patients") %>%
  analyze_num_patients(
    vars = "USUBJID",
    .stats = c("unique", "nonunique"),
    .labels = c(
      unique = "Total number of patients with at least one adverse event",
      nonunique = "Overall total number of events"
    )
  ) %>%
  split_rows_by(
    "AEBODSYS",
    child_labels = "visible",
    nested = FALSE,
    split_fun = split_fun,
    label_pos = "topleft",
    split_label = obj_label(adae$AEBODSYS)
  ) %>%
  summarize_num_patients(
    var = "USUBJID",
    .stats = c("unique", "nonunique"),
    .labels = c(
      unique = "Total number of patients with at least one adverse event",
      nonunique = "Total number of events"
    )
  ) %>%
  count_occurrences(
    vars = "AEDECOD",
    .indent_mods = -1L
  ) %>%
  append_varlabels(adae, "AEDECOD", indent = 1L)

result <- build_table(lyt, df = adae, alt_counts_df = adsl)

result
```
