---
title: "Demographic Table"
order: 1
---

```{r setup script, include=FALSE, purl=FALSE}
invisible_hook_purl <- function(before, options, ...) {knitr::hook_purl(before, options, ...); NULL}
knitr::knit_hooks$set(purl = invisible_hook_purl)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This guide will show you how pharmaverse packages, along with some from tidyverse, can be used to create a Demographic table, using the `{pharmaverseadam}` `ADSL` data as an input.

## Data preprocessing

Now we will add some pre-processing to create some extra formatted variables ready for display in the table.

```{r preproc}
#| message: false
library(dplyr)

# Create categorical variables, remove scren failures, and assign column labels
adsl <- pharmaverseadam::adsl |>
  filter(!ACTARM %in% "Screen Failure") |>
  mutate(
    SEX = case_when(
      SEX == "M" ~ "Male",
      SEX == "F" ~ "Female"
    ) |>
      factor(),
    AGEGR1 =
      case_when(
        between(AGE, 18, 40) ~ "18-40",
        between(AGE, 41, 64) ~ "41-64",
        AGE > 64 ~ ">=65"
      ) |>
      factor(levels = c("18-40", "41-64", ">=65")
      )
  ) |> 
  labelled::set_variable_labels(AGE = "Age (yr)",
                                AGEGR1 = "Age group",
                                SEX = "Sex",
                                RACE = "Race")
```

## {gtsummary} & {cards}

In the example below, we will use the [{gtsummary}](https://www.danieldsjoberg.com/gtsummary/) and [{cards}](https://github.com/insightsengineering/cards) packages to create a demographics tables.

- The {cards} package creates Analysis Results Datasets (ARDs, which are a part of the emerging [CDSIC Analyis Results Standard](https://www.cdisc.org/standards/foundational/analysis-results-standard)).
- The {gtsummary} utilizes ARDs to create tables.

#### ARD to Table

In the example below, we first build an ARD with the needed summary statistics using {cards}.
Then, we use the ARD to build the demographics table.

```{r gtsummary-table}
#| message: false
library(cards)
library(gtsummary)
theme_gtsummary_compact() # reduce default padding and font size for a gt table

# build the ARD with the needed summary statistics using {cards}
ard <-
  ard_stack(
    adsl, 
    ard_continuous(variables = AGE),
    ard_categorical(variables = c(AGEGR1, SEX, RACE)),
    .by = ACTARM,      # split results by treatment arm
    .missing = TRUE,   # add information about missingness rates
    .attributes = TRUE # optionally include column labels in the ARD
  )

# use the ARD to create a demographics table
tbl_ard_summary(
  cards = ard, 
  by = ACTARM, 
  include = c(AGE, AGEGR1, SEX, RACE),
  type = AGE ~ "continuous2",
  statistic = AGE ~ c("{mean} ({sd})", 
                      "{median} ({p25}, {p75})", 
                      "{min}, {max}"),
  missing = "always"
) |> 
  bold_labels() |> 
  modify_header(all_stat_cols() ~ "**{level}**  \nN = {n}") |> # add Ns to header
  modify_footnote(everything() ~ NA) # remove default footnote
```

#### Table to ARD

One may also build the demographics in the classic way using `tbl_summary()` from a data frame, then extract the ARD from the table object.

```{r gtsummary-ard}
tbl <- adsl |> 
  tbl_summary(
    by = ACTARM, 
    include = c(AGE, AGEGR1, SEX, RACE),
    # display summary stats for AGE on multiple rows
    type = AGE ~ "continuous2",
    statistic = AGE ~ c("{mean} ({sd})", "{median} ({p25}, {p75})", "{min}, {max}"),
    missing = "always"
  ) |> 
  bold_labels() |> 
  modify_footnote(everything() ~ NA) # remove default footnote

gather_ard(tbl)[[1]] |> select(-gts_column) # removing column so ARD fits on page
```

## {rtables} & {tern}

The packages used with a brief description of their purpose are as follows:

* [`{rtables}`](https://insightsengineering.github.io/rtables/): designed to create and display complex tables with R.
* [`{tern}`](https://insightsengineering.github.io/tern/): contains analysis functions to create tables and graphs used for clinical trial reporting.

After installation of packages, the first step is to load our pharmaverse packages and input data. Here, we are going to encode missing entries in a data frame `adsl`.

Note that `{tern}` depends on `{rtables}` so the latter is automatically attached.

```{r rtables-setup, message=FALSE, warning=FALSE, results='hold'}
library(tern)

adsl2 <- adsl %>%
  df_explicit_na()
```

Now we create the demographic table.

```{r rtables-table}
vars <- c("AGE", "AGEGR1", "SEX", "RACE")
var_labels <- c(
  "Age (yr)",
  "Age group",
  "Sex",
  "Race"
)

lyt <- basic_table(show_colcounts = TRUE) %>%
  split_cols_by(var = "ACTARM") %>%
  add_overall_col("All Patients") %>%
  analyze_vars(
    vars = vars,
    var_labels = var_labels
  )

result <- build_table(lyt, adsl2)

result
```
