---
title: "Demographic Table"
order: 1
---

## Introduction

This guide will show you how pharmaverse packages, along with some from tidyverse, can be used to create a Demographic table, using the `{pharmaverseadam}` `ADSL` data as an input.

The packages used with a brief description of their purpose are as follows:

* [`{rtables}`](https://roche.github.io/rtables/index.html): designed to create and display complex tables with R.
* [`{tern}`](https://insightsengineering.github.io/tern/main/index.html): contains analysis functions to create tables and graphs used for clinical trial reporting.

## Load Data and Required pharmaverse Package

After installation of packages, the first step is to load our pharmaverse packages and input data. Here, we are going to encode missing entries across groups of categorical variables in a data frame `adsl`.

```{r setup, message=FALSE, warning=FALSE, results='hold'}
library(pharmaverseadam)
library(rtables)
library(tern)
library(dplyr)

adsl <- adsl %>%
  df_explicit_na()

# Create categorical variables
adsl <- adsl %>%
  mutate(
    SEX = factor(case_when(
      SEX == "M" ~ "Male",
      SEX == "F" ~ "Female",
      SEX == "U" ~ "Unknown",
      SEX == "UNDIFFERENTIATED" ~ "Undifferentiated"
    )),
    AGEGR1 = factor(
      case_when(
        between(AGE, 18, 40) ~ "18-40",
        between(AGE, 41, 64) ~ "41-64",
        AGE > 64 ~ ">=65"
      ),
      levels = c("18-40", "41-64", ">=65")
    ),
    BMRKR1_CAT = factor(
      case_when(
        BMRKR1 < 3.5 ~ "LOW",
        BMRKR1 >= 3.5 & BMRKR1 < 10 ~ "MEDIUM",
        BMRKR1 >= 10 ~ "HIGH"
      ),
      levels = c("LOW", "MEDIUM", "HIGH")
    )
  )
```

## Start preprocessing

Now we will add some pre-processing to create some extra formatted variables ready for display in the table.

```{r preproc}
# Create categorical variables
adsl <- adsl %>%
  mutate(
    SEX = factor(case_when(
      SEX == "M" ~ "Male",
      SEX == "F" ~ "Female",
      SEX == "U" ~ "Unknown",
      SEX == "UNDIFFERENTIATED" ~ "Undifferentiated"
    )),
    AGEGR1 = factor(
      case_when(
        between(AGE, 18, 40) ~ "18-40",
        between(AGE, 41, 64) ~ "41-64",
        AGE > 64 ~ ">=65"
      ),
      levels = c("18-40", "41-64", ">=65")
    ),
    BMRKR1_CAT = factor(
      case_when(
        BMRKR1 < 3.5 ~ "LOW",
        BMRKR1 >= 3.5 & BMRKR1 < 10 ~ "MEDIUM",
        BMRKR1 >= 10 ~ "HIGH"
      ),
      levels = c("LOW", "MEDIUM", "HIGH")
    )
  )
```

## Demographic table

Now we create the demographic table.

```{r table}
vars <- c("AGE", "AGEGR1", "SEX", "RACE", "BMRKR1", "BMRKR1_CAT")
var_labels <- c(
  "Age (yr)",
  "Age group",
  "Sex",
  "Race",
  "Continous Level Biomarker 1",
  "Categorical Level Biomarker 1"
)

lyt <- basic_table(show_colcounts = TRUE) %>%
  split_cols_by(var = "ACTARM") %>%
  add_overall_col("All Patients") %>%
  analyze_vars(
    vars = vars,
    var_labels = var_labels
  )

result <- build_table(lyt, adsl)

result
```
