---
title: "Demographic Table"
order: 1
---

```{r setup script, include=FALSE, purl=FALSE}
invisible_hook_purl <- function(before, options, ...) {
  knitr::hook_purl(before, options, ...)
  NULL
}
knitr::knit_hooks$set(purl = invisible_hook_purl)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This guide will show you how pharmaverse packages, along with some from tidyverse, can be used to create a Demographic table, using the `{pharmaverseadam}` `ADSL` data as an input.

In the examples below, we illustrate three general approaches for creating a demographics table.
The first and third utilize Analysis Results Datasets---part of the emerging [CDISC Analysis Results Standard](https://www.cdisc.org/standards/foundational/analysis-results-standard).
The second is the classic method of creating summary tables directly from a data set.

## Data preprocessing

Now we will add some pre-processing to create some extra formatted variables ready for display in the table.

```{r preproc}
#| message: false
library(dplyr)

# Create categorical variables, remove screen failures, and assign column labels
adsl <- pharmaverseadam::adsl |>
  filter(!ACTARM %in% "Screen Failure") |>
  mutate(
    SEX = case_match(SEX, "M" ~ "MALE", "F" ~ "FEMALE"),
    AGEGR1 =
      case_when(
        between(AGE, 18, 40) ~ "18-40",
        between(AGE, 41, 64) ~ "41-64",
        AGE > 64 ~ ">=65"
      ) |>
        factor(levels = c("18-40", "41-64", ">=65"))
  ) |>
  labelled::set_variable_labels(
    AGE = "Age (yr)",
    AGEGR1 = "Age group",
    SEX = "Sex",
    RACE = "Race"
  )
```

## {gtsummary} & {cards}

In the example below, we will use the [{gtsummary}](https://www.danieldsjoberg.com/gtsummary/) and [{cards}](https://insightsengineering.github.io/cards/) packages to create a demographics tables.

- The {cards} package creates Analysis Results Datasets (ARDs, which are a part of the [CDISC Analysis Results Standard](https://www.cdisc.org/standards/foundational/analysis-results-standard)).
- The {gtsummary} utilizes ARDs to create tables.

#### ARD ➡ Table

In the example below, we first build an ARD with the needed summary statistics using {cards}.
Then, we use the ARD to build the demographics table with {gtsummary}.

```{r gtsummary-table}
#| message: false
library(cards)
library(gtsummary)
theme_gtsummary_compact() # reduce default padding and font size for a gt table

# build the ARD with the needed summary statistics using {cards}
ard <-
  ard_stack(
    adsl,
    ard_continuous(variables = AGE),
    ard_categorical(variables = c(AGEGR1, SEX, RACE)),
    .by = ACTARM, # split results by treatment arm
    .attributes = TRUE # optionally include column labels in the ARD
  )

# use the ARD to create a demographics table using {gtsummary}
tbl_ard_summary(
  cards = ard,
  by = ACTARM,
  include = c(AGE, AGEGR1, SEX, RACE),
  type = AGE ~ "continuous2",
  statistic = AGE ~ c("{N}", "{mean} ({sd})", "{median} ({p25}, {p75})", "{min}, {max}")
) |>
  bold_labels() |>
  modify_header(all_stat_cols() ~ "**{level}**  \nN = {n}") |> # add Ns to header
  modify_footnote(everything() ~ NA) # remove default footnote
```

#### Table ➡ ARD

One may also build the demographics in the classic way using `gtsummary::tbl_summary()` from a data frame, then extract the ARD from the table object.

```{r gtsummary-ard}
# build demographics table directly from a data frame
tbl <- adsl |> tbl_summary(by = ACTARM, include = c(AGE, AGEGR1, SEX, RACE))

# extract ARD from table object
gather_ard(tbl)[[1]] |> select(-gts_column) # removing column so ARD fits on page
```

## {rtables} & {tern}

The packages used with a brief description of their purpose are as follows:

* [`{rtables}`](https://insightsengineering.github.io/rtables/): designed to create and display complex tables with R.
* [`{tern}`](https://insightsengineering.github.io/tern/): contains analysis functions to create tables and graphs used for clinical trial reporting.

After installation of packages, the first step is to load our pharmaverse packages and input data. Here, we are going to encode missing entries in a data frame `adsl`.

Note that `{tern}` depends on `{rtables}` so the latter is automatically attached.

```{r rtables-setup, message=FALSE, warning=FALSE, results='hold'}
library(tern)

adsl2 <- adsl |>
  df_explicit_na()
```

Now we create the demographic table.

```{r rtables-table}
vars <- c("AGE", "AGEGR1", "SEX", "RACE")
var_labels <- c(
  "Age (yr)",
  "Age group",
  "Sex",
  "Race"
)

lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by(var = "ACTARM") |>
  add_overall_col("All Patients") |>
  analyze_vars(
    vars = vars,
    var_labels = var_labels
  )

result <- build_table(lyt, adsl2)

result
```

## {tfrmt} & {cards}

In the example below, we will use the [{tfrmt}](https://github.com/GSK-Biostatistics/tfrmt) and [{cards}](https://insightsengineering.github.io/cards/) packages to create a demographics tables.

- The {cards} package creates Analysis Results Datasets (ARDs, which are a part of the [CDISC Analysis Results Standard](https://www.cdisc.org/standards/foundational/analysis-results-standard)).
- The {tfrmt} utilizes ARDs to create tables.

In the example below, we first build an ARD with the needed summary statistics using {cards}.
Then, we use the ARD to build the demographics table with {tfrmt}.

```{r tfrmt-table}
#| message: false
library(cards)
library(forcats)
library(tfrmt)

# build the ARD with the needed summary statistics using {cards}
ard <-
  ard_stack(
    adsl,
    ard_continuous(variables = AGE,
                   statistic = ~ continuous_summary_fns(c("N", "mean", "sd", "min", "max"))),
    ard_categorical(variables = c(AGEGR1, SEX, RACE)),
    .by = ACTARM, # split results by treatment arm
    .overall = TRUE,
    .total_n = TRUE
  )

# tidy the ARD for use in {tfrmt}
ard_tbl <-
  ard |>
  #reshape the data
  shuffle_card(fill_overall = "Total") |>
  
  # transform group-level freqs/pcts into a singular "bigN" row
  prep_big_n(vars = "ACTARM") |> 
  
  # consolidate vars into a single variable column
  prep_combine_vars(vars = c("AGE","AGEGR1","SEX","RACE")) |> 
  
  # coalesce categorical levels + continuous stats into a "label"
  prep_label() |> 
  
  group_by(ACTARM, stat_variable) |> 
  mutate(across(c(variable_level, label), ~ ifelse(stat_name=="N", "n", .x))) |> 
  ungroup() |> 
  unique() |> 
  # sorting 
  mutate(
    ord1 = fct_inorder(stat_variable) |> fct_relevel("SEX", after = 0) |> as.numeric(),
    ord2 = ifelse(label == "n", 1, 2)
  ) |>
  # relabel the variables
  mutate(stat_variable = case_when(
    stat_variable == "AGE" ~ "Age (YEARS) at First Dose",
    stat_variable == "AGEGR1" ~ "Age Group (YEARS) at First Dose",
    stat_variable == "SEX" ~ "Sex",
    stat_variable == "RACE" ~ "High Level Race",
    .default = stat_variable
  )) |>
  # drop variables not needed
  select(ACTARM, stat_variable, label, stat_name, stat, ord1, ord2) |>
  # remove dups (extra denoms per variable level)
  unique()

# create a demographics table using {tfrmt}
DM_T01 <- tfrmt(
  group = stat_variable,
  label = label,
  param = stat_name,
  value = stat,
  column = ACTARM,
  sorting_cols = c(ord1, ord2),
  body_plan = body_plan(
    frmt_structure(group_val = ".default", label_val = ".default", frmt("xxx")),
    frmt_structure(
      group_val = ".default", label_val = ".default",
      frmt_combine("{n} ({p}%)",
        n = frmt("xxx"),
        p = frmt("xx", transform = ~ . * 100)
      )
    )
  ),
  big_n = big_n_structure(param_val = "bigN", n_frmt = frmt(" (N=xx)")),
  col_plan = col_plan(
    -starts_with("ord")
  ),
  col_style_plan = col_style_plan(
      col_style_structure(col = c("Placebo", "Xanomeline High Dose", "Xanomeline Low Dose", "Total"), align = "left")
    ),
  row_grp_plan = row_grp_plan(
    row_grp_structure(group_val = ".default", element_block(post_space = " "))
  )
) |>
  print_to_gt(ard_tbl)

DM_T01

```