---
title: "Oncology Survival"
order: 4
---

```{r setup script, include=FALSE, purl=FALSE}
invisible_hook_purl <- function(before, options, ...) {
  knitr::hook_purl(before, options, ...)
  NULL
}
knitr::knit_hooks$set(purl = invisible_hook_purl)
```

## Introduction

This guide demonstrates how pharmaverse packages, along with tools from the
tidyverse, can be used to create standard oncology survival Tables, Listings,
and Graphs (TLGs) using the `{pharmaverseadam}` `ADTTE` dataset as input.

The packages used, with a brief description of their purpose, are as follows:

* [`{pharmaverseadam}`](https://pharmaverse.github.io/pharmaverseadam/): provides
  CDISC ADaM example datasets for use in pharmaverse examples and tests.
* [`{ggsurvfit}`](https://www.danieldsjoberg.com/ggsurvfit/): eases the creation
  of publication-ready time-to-event (survival) figures built on `{ggplot2}`.
  Includes `Surv_CNSR()` to handle CDISC ADTTE censoring conventions natively.
* [`{gtsummary}`](https://www.danieldsjoberg.com/gtsummary/): creates
  publication-ready summary and analytical tables. Works seamlessly with
  `{ggsurvfit}` to generate median survival tables.
* [`{dplyr}`](https://dplyr.tidyverse.org/): provides data manipulation
  functions used to prepare and filter the ADTTE data.

The outputs produced in this example are:

1. **Kaplan-Meier plot** with confidence intervals, risk table, and median
   guideline — a standard figure for oncology efficacy reporting.
2. **Risk table** displayed beneath the KM plot showing number at risk, events,
   and censored counts at each time point.
3. **Median survival table** with 95% confidence interval, suitable for
   inclusion in a clinical study report (CSR).

---

## Setup

We load the required packages and read `ADTTE` from `{pharmaverseadam}`. Because
the CDISC ADTTE censoring variable `CNSR` is coded `1 = censored / 0 = event`
(the reverse of base R's `survival::Surv()` convention), we use `Surv_CNSR()`
from `{ggsurvfit}` throughout to avoid any mis-coding.

```{r setup, message=FALSE, warning=FALSE}
library(pharmaverseadam)
library(ggsurvfit)
library(gtsummary)
library(dplyr)

# ── Read data ──────────────────────────────────────────────────────────────────
adtte <- pharmaverseadam::adtte_onco

# Keep a single endpoint for the basic example: Overall Survival (OS).
# Adjust PARAMCD to the endpoint of interest for other analyses.
adtte_os <- adtte |>
  filter(PARAMCD == "OS")

# Preview key variables
adtte_os |>
  select(USUBJID, PARAM, PARAMCD, AVAL, CNSR, EVNTDESC, CNSDTDSC) |>
  head(5)
```

> **Note on CNSR coding:** In CDISC ADTTE, `CNSR = 0` indicates an **event**
> and `CNSR = 1` indicates **censoring**. `Surv_CNSR(AVAL, CNSR)` handles this
> correctly without requiring manual recoding.

---

## Kaplan-Meier Plot

### Basic KM Curve with Risk Table

The plot below uses `survfit2()` with `Surv_CNSR()` to fit the survival model.
Because `PARAM` and `PARAMCD` are present in the data, `{ggsurvfit}`
automatically uses `PARAM` as the y-axis label — a key CDISC-aware feature.

```{r km-plot, message=FALSE, warning=FALSE, fig.width=9, fig.height=7}
# ── Fit the survival model ─────────────────────────────────────────────────────
# survfit2() is preferred over survfit() as it tracks the calling environment,
# enabling clean legend labels and p-value computation downstream.
km_fit <- survfit2(Surv_CNSR(AVAL, CNSR) ~ 1, data = adtte_os)

# ── Build the plot ─────────────────────────────────────────────────────────────
km_fit |>
  ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_risktable(
    risktable_stats = c("n.risk", "cum.event", "cum.censor"),
    stats_label = list(
      n.risk = "At Risk",
      cum.event = "Events (cum.)",
      cum.censor = "Censored (cum.)"
    )
  ) +
  add_quantile(
    y_value   = 0.5, # median survival guideline
    color     = "gray40",
    linewidth = 0.75,
    linetype  = "dashed"
  ) +
  add_censor_mark(shape = 3, size = 2) +
  scale_ggsurvfit(x_scales = list(breaks = 0:6)) +
  labs(
    title = paste0(unique(adtte_os$PARAM), "\nKaplan-Meier Estimate"),
    x = "Time (Years)",
    y = "Overall Survival Probability",
    caption = paste0(
      "Analysis dataset: ADTTE  |  PARAMCD: ", unique(adtte_os$PARAMCD),
      "\nCensored observations marked with '+'"
    )
  ) +
  theme_ggsurvfit_default() +
  theme(plot.caption = element_text(hjust = 0, size = 8))
```

**Figure notes:**

* The shaded band shows the pointwise 95% confidence interval (default:
  log-log transformed).
* The dashed horizontal line marks the 50% survival probability; its
  intersection with the curve gives the median survival time.
* `+` marks indicate individual censored observations.
* The risk table beneath the plot shows cumulative counts aligned to the
  x-axis breaks.

---

## Median Survival Table

`{gtsummary}`'s `tbl_survfit()` function extracts summary statistics from a
`survfit` object and formats them as a publication-ready table. It respects
CDISC ADTTE variable names when using `Surv_CNSR()`.

```{r median-table, message=FALSE, warning=FALSE}
# ── Median survival with 95% CI ────────────────────────────────────────────────
tbl_survfit(
  km_fit,
  probs = 0.5, # 50th percentile = median
  label_header = "**Median (95% CI)**"
) |>
  modify_caption(
    paste0(
      "**Table 1. Median Overall Survival**",
      "\nADTTE  |  PARAMCD: ", unique(adtte_os$PARAMCD)
    )
  ) |>
  bold_labels()
```

---

## Survival Probability Table

This table shows estimated survival probabilities at clinically meaningful
time points (6, 12, 18, and 24 months = 0.5, 1, 1.5, 2 years), which are
commonly reported in oncology CSR summaries.

```{r prob-table, message=FALSE, warning=FALSE}
# ── Survival probability at selected time points ───────────────────────────────
tbl_survfit(
  km_fit,
  times        = c(0.5, 1, 1.5, 2), # time in years (AVAL units in adtte)
  label_header = "**Survival Probability (95% CI)**"
) |>
  modify_header(
    label = "**Time Point (Years)**"
  ) |>
  modify_caption(
    paste0(
      "**Table 2. Overall Survival Probability at Selected Time Points**",
      "\nADTTE  |  PARAMCD: ", unique(adtte_os$PARAMCD)
    )
  ) |>
  bold_labels()
```

---

## Key Package Notes

**`Surv_CNSR()` vs `Surv()`**

CDISC ADTTE datasets use `CNSR = 0` for events and `CNSR = 1` for censoring —
the opposite of `survival::Surv()`. Using `Surv_CNSR(AVAL, CNSR)` removes
this error-prone manual recoding step:

```{r cnsr-note, eval=FALSE}
# ✗  Error-prone: requires manual recoding of CNSR
survival::Surv(adtte_os$AVAL, 1 - adtte_os$CNSR)

# ✓  Correct CDISC-aware approach
ggsurvfit::Surv_CNSR(adtte_os$AVAL, adtte_os$CNSR)
```

**`survfit2()` vs `survfit()`**

`survfit2()` tracks the calling environment, which enables `{ggsurvfit}` to
cleanly remove raw variable names (e.g. `CNSR=0`) from figure legends and to
compute p-values via `survfit2_p()`. For ADTTE data, it also reads `PARAM`/`PARAMCD`
automatically to populate axis labels.

**Extending to multiple strata**

To compare arms (e.g. `TRT01A`), change the formula and add arm-level variables
from `ADSL` via a merge:

```{r strata-note, eval=FALSE}
# Merge treatment from ADSL
adsl <- pharmaverseadam::adsl
adtte_os_arm <- adtte_os |>
  left_join(select(adsl, USUBJID, TRT01A), by = "USUBJID")

# Stratified KM
survfit2(Surv_CNSR(AVAL, CNSR) ~ TRT01A, data = adtte_os_arm) |>
  ggsurvfit(linewidth = 1) +
  add_confidence_interval() +
  add_risktable() +
  scale_ggsurvfit()
```
