---
title: "Oncology Survival"
order: 4
---

```{r setup script, include=FALSE, purl=FALSE}
invisible_hook_purl <- function(before, options, ...) {
  knitr::hook_purl(before, options, ...)
  NULL
}
knitr::knit_hooks$set(purl = invisible_hook_purl)
```

## Introduction

This guide demonstrates how pharmaverse packages, along with tools from the
tidyverse, can be used to create standard oncology survival Tables, Listings,
and Graphs (TLGs) using the `{pharmaverseadam}` `ADTTE_ONCO` dataset as input.

The packages used, with a brief description of their purpose, are as follows:

* [`{pharmaverseadam}`](https://pharmaverse.github.io/pharmaverseadam/): provides
  CDISC ADaM example datasets for use in pharmaverse examples and tests.
* [`{ggsurvfit}`](https://www.danieldsjoberg.com/ggsurvfit/): eases the creation
  of publication-ready time-to-event (survival) figures built on `{ggplot2}`.
  Includes `Surv_CNSR()` to handle CDISC ADTTE censoring conventions natively.
* [`{gtsummary}`](https://www.danieldsjoberg.com/gtsummary/): creates
  publication-ready summary and analytical tables. Works seamlessly with
  `{ggsurvfit}` to generate median survival tables.
* [`{dplyr}`](https://dplyr.tidyverse.org/): provides data manipulation
  functions used to prepare and filter the ADTTE data.

The outputs produced in this example are:

1. **Kaplan-Meier plot** with confidence intervals, risk table, and median
   guideline — a standard figure for oncology efficacy reporting.
2. **Risk table** displayed beneath the KM plot showing number at risk, events,
   and censored counts at each time point.
3. **Median survival table** with 95% confidence interval, suitable for
   inclusion in a clinical study report (CSR).
4. **Survival probability table** at selected time points.
5. **Stratified KM plot** using the built-in `ggsurvfit::adtte` four-arm trial
   dataset to illustrate multi-arm analyses.

---

## Setup

We load the required packages and read `ADTTE_ONCO` from `{pharmaverseadam}`.
This is the oncology-specific TTE dataset generated from the `{admiralonco}`
template. It contains three endpoints — Overall Survival (`OS`), Progression
Free Survival (`PFS`), and Duration of Response (`RSD`) — with treatment arm
variables (`ARM`, `ARMCD`) already merged in, making it ready for stratified
analyses without a separate ADSL join.

Because the CDISC ADTTE censoring variable `CNSR` is coded `1 = censored /
0 = event` (the reverse of base R's `survival::Surv()` convention), we use
`Surv_CNSR()` from `{ggsurvfit}` throughout to avoid mis-coding errors.

```{r setup}
#| message: false
#| warning: false
library(pharmaverseadam)
library(ggsurvfit)
library(ggplot2)
library(gtsummary)
library(dplyr)

# ── Read data ──────────────────────────────────────────────────────────────────
adtte_onco <- pharmaverseadam::adtte_onco

# Overview of available endpoints and their event rates
adtte_onco |>
  group_by(PARAMCD, PARAM) |>
  summarise(
    N           = n(),
    N_events    = sum(CNSR == 0),
    Pct_events  = round(100 * mean(CNSR == 0), 1),
    Median_AVAL = round(median(AVAL), 2),
    .groups     = "drop"
  )
```

> **Why PFS for this example?** OS in early-phase oncology trials typically has
> few events (most subjects are still alive), producing a near-flat KM curve
> with a median that cannot be estimated — not ideal for illustration. PFS has
> more events and a shorter follow-up window, giving a more informative curve.
> The code below uses `PARAMCD == "PFS"`; swap to `"OS"` or `"RSD"` as needed.
> Note that `{admiralonco}` uses `"RSD"` (not `"DOR"`) for Duration of Response.

```{r filter-pfs}
#| message: false
#| warning: false

# ── Filter to PFS endpoint ─────────────────────────────────────────────────────
adtte_pfs <- adtte_onco |>
  filter(PARAMCD == "PFS")

# Preview key variables
adtte_pfs |>
  select(USUBJID, PARAM, PARAMCD, AVAL, CNSR, EVNTDESC, CNSDTDSC) |>
  head(5)
```

> **Note on CNSR coding:** In CDISC ADTTE, `CNSR = 0` indicates an **event**
> and `CNSR = 1` indicates **censoring**. `Surv_CNSR(AVAL, CNSR)` handles this
> correctly without requiring manual recoding.

---

## Kaplan-Meier Plot

The plot below uses `survfit2()` with `Surv_CNSR()` to fit the survival model.
Because `PARAM` and `PARAMCD` are present in the data, `{ggsurvfit}`
automatically uses `PARAM` as the y-axis label — a key CDISC-aware feature.
The KM plot is the same whether you use the Classic or ARD approach for the
tables, so it is shown once here.

```{r km-plot}
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 7

# ── Fit the survival model ─────────────────────────────────────────────────────
# survfit2() is preferred over survfit() as it tracks the calling environment,
# enabling clean legend labels and p-value computation downstream.
km_fit <- survfit2(Surv_CNSR(AVAL, CNSR) ~ 1, data = adtte_pfs)

# ── Build the plot ─────────────────────────────────────────────────────────────
km_fit |>
  ggsurvfit(linewidth = 1) +
  # ── Explicit color scales prevent the site theme from stripping colour to B&W ─
  scale_color_manual(values = c("#2c7bb6")) +
  scale_fill_manual(values = c("#2c7bb6")) +
  add_confidence_interval() +
  add_risktable(
    risktable_stats = c("n.risk", "cum.event", "cum.censor"),
    stats_label = list(
      n.risk     = "At Risk",
      cum.event  = "Events (cum.)",
      cum.censor = "Censored (cum.)"
    ),
    # ── Slightly smaller label text keeps row labels from squeezing number columns
    theme = theme_risktable_default(axis.text.y.size = 9, plot.title.size = 9)
  ) +
  add_quantile(
    y_value   = 0.5, # median survival guideline
    color     = "gray40",
    linewidth = 0.75,
    linetype  = "dashed"
  ) +
  add_censor_mark(shape = 3, size = 2) +
  scale_ggsurvfit() +
  labs(
    title = paste0(unique(adtte_pfs$PARAM), "\nKaplan-Meier Estimate"),
    x = paste0("Time (", unique(adtte_pfs$AVALU), ")"),
    y = "Progression-Free Survival Probability",
    caption = paste0(
      "Analysis dataset: ADTTE_ONCO  |  PARAMCD: ", unique(adtte_pfs$PARAMCD),
      "\nCensored observations marked with '+'"
    )
  ) +
  theme_ggsurvfit_default() +
  theme(plot.caption = element_text(hjust = 0, size = 8))
```

**Figure notes:**

* The shaded band shows the pointwise 95% confidence interval (default:
  log-log transformed).
* The dashed horizontal line marks the 50% survival probability; its
  intersection with the curve gives the median survival time.
* `+` marks indicate individual censored observations.
* The risk table beneath the plot shows cumulative counts aligned to the
  x-axis breaks.

---

## Classic Approach

In the Classic approach, `{gtsummary}`'s `tbl_survfit()` consumes the `km_fit`
`survfit2` object directly to produce publication-ready tables. No intermediate
ARD layer is created; the statistics are extracted and formatted in a single
pipeline.

### Median Survival Table

```{r median-table-classic}
#| message: false
#| warning: false

# ── Median survival with 95% CI ────────────────────────────────────────────────
tbl_survfit(
  km_fit,
  probs        = 0.5, # 50th percentile = median
  label_header = "**Median (95% CI)**"
) |>
  modify_caption(
    paste0(
      "**Table 1. Median Progression-Free Survival**",
      "\nADTTE_ONCO  |  PARAMCD: ", unique(adtte_pfs$PARAMCD)
    )
  ) |>
  bold_labels()
```

### Survival Probability Table

This table shows estimated PFS probabilities at clinically meaningful time
points. For PFS these are typically shorter intervals than OS — 3, 6, 9, and
12 months are standard in many oncology trials. Adjust `times` to match the
`AVALU` units in your dataset.

```{r prob-table-classic}
#| message: false
#| warning: false

# ── Survival probability at selected time points ───────────────────────────────
# AVAL in adtte_onco is in years; express months as fractions of a year
tbl_survfit(
  km_fit,
  times        = c(0.25, 0.5, 0.75, 1), # 3, 6, 9, 12 months in years
  label_header = "**PFS Probability (95% CI)**"
) |>
  modify_header(
    label = "**Time Point**"
  ) |>
  modify_table_body(
    ~ .x |>
      mutate(label = recode(label,
        "0.25" = "3 months",
        "0.5"  = "6 months",
        "0.75" = "9 months",
        "1"    = "12 months"
      ))
  ) |>
  modify_caption(
    paste0(
      "**Table 2. Progression-Free Survival Probability at Selected Time Points**",
      "\nADTTE_ONCO  |  PARAMCD: ", unique(adtte_pfs$PARAMCD)
    )
  ) |>
  bold_labels()
```


::: {.callout-note}
## ARD output

Every `{gtsummary}` table stores its underlying statistics in a `.$cards` slot
as an Analysis Results Data (ARD) object conforming to the emerging CDISC
Analysis Results Standard. You can also generate a standalone survival ARD for
audit, downstream rendering, or cross-validation using
`cardx::ard_survival_survfit()`:

```r

# Median survival ARD
adtte_pfs |>
  cardx::ard_survival_survfit(
    y     = "Surv_CNSR(AVAL, CNSR)",
    probs = 0.5
  )

# Time-point survival probability ARD
adtte_pfs |>
  cardx::ard_survival_survfit(
    y     = "Surv_CNSR(AVAL, CNSR)",
    times = c(0.25, 0.5, 0.75, 1)
  )
```

The `y` argument must be passed as a character string when using the data frame
method. See the [`{cardx}` documentation](https://insightsengineering.github.io/cardx/main/reference/ard_survival_survfit.html)
for full details.
:::

---

## Key Package Notes

**`Surv_CNSR()` vs `Surv()`**

CDISC ADTTE datasets use `CNSR = 0` for events and `CNSR = 1` for censoring —
the opposite of `survival::Surv()`. Using `Surv_CNSR(AVAL, CNSR)` removes
this error-prone manual recoding step, and it works identically in both the
`survfit2()` call for plotting and the `ard_survival_survfit()` call for ARDs:

```{r cnsr-note}
#| eval: false

# ✗  Error-prone: requires manual recoding of CNSR
survival::Surv(adtte_pfs$AVAL, 1 - adtte_pfs$CNSR)

# ✓  Correct CDISC-aware approach
ggsurvfit::Surv_CNSR(adtte_pfs$AVAL, adtte_pfs$CNSR)
```

**`survfit2()` vs `survfit()`**

`survfit2()` tracks the calling environment, which enables `{ggsurvfit}` to
cleanly remove raw variable names (e.g. `CNSR=0`) from figure legends and to
compute p-values via `survfit2_p()`. For ADTTE data, it also reads `PARAM`/`PARAMCD`
automatically to populate axis labels. Use `survfit2()` for all plotting work.

**Switching endpoints**

All outputs above are driven by the `adtte_pfs` object. To switch
to OS or RSD, simply change the filter at the top of the Setup section:

```{r endpoint-note}
#| eval: false

# Overall Survival — expect few events; median may not be estimable
adtte_os <- adtte_onco |> filter(PARAMCD == "OS")

# Duration of Response — responders only; smaller N than OS/PFS
# Note: admiralonco uses PARAMCD = "RSD", not "DOR"
adtte_rsd <- adtte_onco |> filter(PARAMCD == "RSD")
```

**Extending to multiple strata**

Because `adtte_onco` already contains `ARM` and `ARMCD` from the `{admiralonco}`
template, no ADSL join is needed to produce a stratified KM plot:

```{r strata-note}
#| eval: false

survfit2(Surv_CNSR(AVAL, CNSR) ~ ARM, data = adtte_pfs) |>
  ggsurvfit(linewidth = 1) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  add_confidence_interval() +
  add_risktable(
    theme = theme_risktable_default(axis.text.y.size = 9, plot.title.size = 9)
  ) +
  add_pvalue(location = "annotation") +
  scale_ggsurvfit()
```
---

## Stratified KM Plot: `ggsurvfit::adtte`

`{ggsurvfit}` ships its own example ADTTE dataset — a four-arm Phase III breast
cancer trial (2,199 subjects, PFS endpoint, ~34% event rate) originally from the
[VIS-SIG Wonderful Wednesdays](https://github.com/VIS-SIG/Wonderful-Wednesdays/tree/master/data/2020/2020-04-08)
initiative. It provides a richer illustrative dataset than `adtte_onco` for
showing a **stratified KM plot**, since it has four well-separated treatment arms
with a good event rate and an estimable median.

The treatment arms are stored in `STR01` (not `ARM` or `TRT01A`) and have short
coded values (`V`, `T`, `T->V`, `T+V`). Mapping these to readable labels in the
`scale_color_*` call makes the legend self-explanatory.

```{r km-plot-adtte}
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 8

# ── ggsurvfit::adtte — four-arm breast cancer PFS trial ────────────────────────
arm_labels <- c(
  "V"    = "Vinorelbine (V)",
  "T"    = "Docetaxel (T)",
  "T->V" = "Docetaxel → Vinorelbine",
  "T+V"  = "Docetaxel + Vinorelbine"
)

survfit2(Surv_CNSR(AVAL, CNSR) ~ STR01, data = ggsurvfit::adtte) |>
  ggsurvfit(linewidth = 1) +
  scale_color_brewer(palette = "Dark2", labels = arm_labels) +
  scale_fill_brewer(palette = "Dark2", labels = arm_labels) +
  add_confidence_interval() +
  add_risktable(
    risktable_stats = "n.risk",
    stats_label = list(n.risk = "At Risk"),
    theme = theme_risktable_default(axis.text.y.size = 9, plot.title.size = 9)
  ) +
  add_quantile(
    y_value   = 0.5,
    color     = "gray40",
    linewidth = 0.75,
    linetype  = "dashed"
  ) +
  add_censor_mark(shape = 3, size = 1.5) +
  add_pvalue(location = "annotation", x = 4.5) +
  scale_ggsurvfit() +
  labs(
    title = "Progression-Free Survival by Treatment Arm",
    x = "Time (Years)",
    y = "Progression-Free Survival Probability",
    caption = paste0(
      "Dataset: ggsurvfit::adtte  |  HER2+ breast cancer Phase III trial",
      "\nCensored observations marked with '+'"
    )
  ) +
  theme_ggsurvfit_default() +
  theme(
    plot.caption  = element_text(hjust = 0, size = 8),
    legend.title  = element_blank()
  )
```

**Figure notes:**

* The risk table shows only `n.risk` here (rather than cumulative events and
  censored) to keep the four-strata table readable at standard figure width.
* The log-rank p-value is placed as an annotation rather than in the caption
  to keep it visible when the figure is used standalone.
* `ggsurvfit::adtte` contains only a single PFS endpoint per subject, so no
  `filter(PARAMCD == ...)` step is needed.
