---
title: "The Difference Between `logr`, `logrx`, and `whirl`"
format:
  html:
    code-overflow: wrap
---

```{css, echo=FALSE}
.output-block {
  max-height: 400px;
  overflow-y: scroll;
}
```

The [`logr`](https://github.com/dbosak01/logr/) package for R aims to provide a straightforward logging system tailored for everyday users, such as statisticians and educators, emphasizing simplicity in recording program execution. In contrast, [`logrx`](https://github.com/pharmaverse/logrx/) offers an advanced logging mechanism designed specifically for clinical reporting, featuring easy traceability and reproducibility of R script executions with capabilities suited for more technical users.
[`whirl`](https://github.com/NovoNordisk-OpenSource/whirl)[^1] differs in that the package provides functionalities for executing scripts in batch and simultaneously getting a log from the individual executions. In contrast the other two logging packages, the default logs are HTML reports making it easier to read, and they include the results printed in the programs being logged.

All three packages are used for keeping a record of what happens when R scripts are run, but they differ in what they offer and who they're for. When choosing between them, you should consider what you need for your particular situation.

[^1]: Note: `whirl` is currently only available on GitHub, and not on CRAN, but the development team is working on getting it published shortly.

### Execution

Each package differs notably in the approach to generating a log file. The `logr` package adopts a method similar to SAS, where the log is manually constructed at intermediary steps within the R script. Conversely, the `logrx` package allows for the creation of a log file through a single command from the R terminal, offering added compatibility for command line submission. `whirl` also provides a single function to create log files, and it should be noted that the execution is done through [Quarto](https://quarto.org), which makes enables the HTML reports, but also make it more heavy on external dependencies.

::: {.panel-tabset}

## `logr`

```{r, eval=FALSE}
#| class: output-block
# file: script.R
library(dplyr)
library(logr)

file_path <- "script.log"
log_file <- log_open(file_path)

log_print("Analyzing Fuel Efficiency Trends in Cars")

average_mpg <- mtcars %>%
  group_by(cyl) %>%
  summarise(average_mpg = mean(mpg, na.rm = TRUE))

log_print(average_mpg)

model <- lm(mpg ~ wt, data = mtcars)
summary(model)

log_print(model)

log_close() # log file path: log/script.log
```

<br>

```{t}
#| class: output-block
=========================================================================
Log Path: ./log/script.log
Working Directory: /Users/davidblair/My Drive/Appsilon/Work Projects/examples
User Name: davidblair
R Version: 4.3.3 (2024-02-29)
Machine: Davids-Air.localdomain arm64
Operating System: Darwin 23.4.0 Darwin Kernel Version 23.4.0: Fri Mar 15 00:12:41 PDT 2024; root:xnu-10063.101.17~1/RELEASE_ARM64_T8103
Base Packages: stats graphics grDevices utils datasets methods base
Other Packages: dplyr_1.1.4 logr_1.3.6
Log Start Time: 2024-04-08 13:25:53.22436
=========================================================================

Analyzing Fuel Efficiency Trends in Cars

NOTE: Log Print Time:  2024-04-08 13:25:53.226347
NOTE: Elapsed Time: 0.00139403343200684 secs

# A tibble: 3 × 2
    cyl average_mpg
  <dbl>       <dbl>
1     4        26.7
2     6        19.7
3     8        15.1

NOTE: Data frame has 3 rows and 2 columns.

NOTE: Log Print Time:  2024-04-08 13:25:53.254377
NOTE: Elapsed Time: 0.0280299186706543 secs


Call:
lm(formula = mpg ~ wt, data = mtcars)

Coefficients:
(Intercept)           wt
     37.285       -5.344


NOTE: Log Print Time:  2024-04-08 13:25:53.267483
NOTE: Elapsed Time: 0.013106107711792 secs

=========================================================================
Log End Time: 2024-04-08 13:25:53.286294
Log Elapsed Time: 0 00:00:00
=========================================================================
```

## `logrx`

From the R terminal:
```{r, eval=FALSE}
logrx::axecute("script.R")
```

From the Command Line:
```{bash, eval=FALSE}
Rscript -e "logrx::axecute('script.R')"
```



```{t}
#| class: output-block
--------------------------------------------------------------------------------
-                                logrx Metadata                                -
--------------------------------------------------------------------------------
This log was generated using logrx 0.3.0
logrx package version: 0.3.0
logrx build: CRAN (R 4.3.1)
logrx link to repository: https://github.com/pharmaverse/logrx
--------------------------------------------------------------------------------
-                          User and File Information                           -
--------------------------------------------------------------------------------
User: davidblair
File Name: script.R
File Path: /Users/davidblair/My Drive/Appsilon/Work Projects/examples/logging
File HashSum: 76f9dd029df25ba4f65de72631fac248fa30d7d5
--------------------------------------------------------------------------------
-                             Session Information                              -
--------------------------------------------------------------------------------
─ Session info ─────────────────────────────────────────────────────────────────
 setting  value
 version  R version 4.3.3 (2024-02-29)
 os       macOS Sonoma 14.4.1
 system   aarch64, darwin20
 ui       X11
 language (EN)
 collate  en_US.UTF-8
 ctype    en_US.UTF-8
 tz       Europe/London
 date     2024-04-08
 pandoc   NA

─ Packages ─────────────────────────────────────────────────────────────────────
 ! package     * version date (UTC) lib source
   cachem        1.0.8   2023-05-01 [1] CRAN (R 4.3.0)
   callr         3.7.6   2024-03-25 [1] CRAN (R 4.3.1)
   cli           3.6.2   2023-12-11 [1] CRAN (R 4.3.1)
   common        1.1.2   2024-03-05 [1] CRAN (R 4.3.1)
   curl          5.2.1   2024-03-01 [1] CRAN (R 4.3.1)
   desc          1.4.3   2023-12-10 [1] CRAN (R 4.3.1)
   devtools      2.4.5   2022-10-11 [1] CRAN (R 4.3.0)
   digest        0.6.35  2024-03-11 [1] CRAN (R 4.3.1)
   dplyr       * 1.1.4   2023-11-17 [1] CRAN (R 4.3.1)
   ellipsis      0.3.2   2021-04-29 [1] CRAN (R 4.3.0)
   fansi         1.0.6   2023-12-08 [1] CRAN (R 4.3.1)
   fastmap       1.1.1   2023-02-24 [1] CRAN (R 4.3.0)
   fs            1.6.3   2023-07-20 [1] CRAN (R 4.3.0)
   generics      0.1.3   2022-07-05 [1] CRAN (R 4.3.0)
   glue          1.7.0   2024-01-09 [1] CRAN (R 4.3.1)
   htmltools     0.5.8.1 2024-04-04 [1] CRAN (R 4.3.1)
   htmlwidgets   1.6.4   2023-12-06 [1] CRAN (R 4.3.1)
   httpuv        1.6.15  2024-03-26 [1] CRAN (R 4.3.1)
   later         1.3.2   2023-12-06 [1] CRAN (R 4.3.1)
   lifecycle     1.0.4   2023-11-07 [1] CRAN (R 4.3.1)
 V logr        * 1.3.7   2024-02-24 [1] CRAN (R 4.3.1) (on disk 1.3.6)
   logrx         0.3.0   2023-10-17 [1] CRAN (R 4.3.1)
   magrittr      2.0.3   2022-03-30 [1] CRAN (R 4.3.0)
   memoise       2.0.1   2021-11-26 [1] CRAN (R 4.3.0)
   mime          0.12    2021-09-28 [1] CRAN (R 4.3.0)
   miniUI        0.1.1.1 2018-05-18 [1] CRAN (R 4.3.0)
   pillar        1.9.0   2023-03-22 [1] CRAN (R 4.3.0)
   pkgbuild      1.4.4   2024-03-17 [1] CRAN (R 4.3.1)
   pkgconfig     2.0.3   2019-09-22 [1] CRAN (R 4.3.0)
   pkgload       1.3.4   2024-01-16 [1] CRAN (R 4.3.1)
   processx      3.8.4   2024-03-16 [1] CRAN (R 4.3.1)
   profvis       0.3.8   2023-05-02 [1] CRAN (R 4.3.0)
   promises      1.2.1   2023-08-10 [1] CRAN (R 4.3.0)
   ps            1.7.6   2024-01-18 [1] CRAN (R 4.3.1)
   purrr         1.0.2   2023-08-10 [1] CRAN (R 4.3.0)
   R6            2.5.1   2021-08-19 [1] CRAN (R 4.3.0)
   Rcpp          1.0.12  2024-01-09 [1] CRAN (R 4.3.1)
   remotes       2.4.2.1 2023-07-18 [1] CRAN (R 4.3.0)
   rlang         1.1.3   2024-01-10 [1] CRAN (R 4.3.1)
   rstudioapi    0.15.0  2023-07-07 [1] CRAN (R 4.3.0)
   sessioninfo   1.2.2   2021-12-06 [1] CRAN (R 4.3.0)
   shiny         1.8.1.1 2024-04-02 [1] CRAN (R 4.3.1)
   stringi       1.8.3   2023-12-11 [1] CRAN (R 4.3.1)
   stringr       1.5.1   2023-11-14 [1] CRAN (R 4.3.1)
   tibble        3.2.1   2023-03-20 [1] CRAN (R 4.3.0)
   tidyr         1.3.1   2024-01-24 [1] CRAN (R 4.3.1)
   tidyselect    1.2.1   2024-03-11 [1] CRAN (R 4.3.1)
   urlchecker    1.0.1   2021-11-30 [1] CRAN (R 4.3.0)
   usethis       2.2.3   2024-02-19 [1] CRAN (R 4.3.1)
   utf8          1.2.4   2023-10-22 [1] CRAN (R 4.3.1)
   vctrs         0.6.5   2023-12-01 [1] CRAN (R 4.3.1)
   waiter        0.2.5   2022-01-03 [1] CRAN (R 4.3.0)
   withr         3.0.0   2024-01-16 [1] CRAN (R 4.3.1)
   xtable        1.8-4   2019-04-21 [1] CRAN (R 4.3.0)

 [1] /Library/Frameworks/R.framework/Versions/4.3-arm64/Resources/library

 V ── Loaded and on-disk version mismatch.

─ External software ────────────────────────────────────────────────────────────
 setting        value
 cairo
 cairoFT
 pango
 png
 jpeg
 tiff
 tcl
 curl           8.4.0
 zlib           1.2.12
 bzlib          1.0.8, 13-Jul-2019
 xz             5.4.4
 PCRE           10.42 2022-12-11
 ICU            74.1
 TRE            TRE 0.8.0 R_fixes (BSD)
 iconv          Apple or GNU libiconv 1.11
 readline       5.2

	  BLAS           /Library/Frameworks/R.framework/Versions/4.3-arm64/Resources/lib/libRblas.0.dylib

	  lapack         /Library/Frameworks/R.framework/Versions/4.3-arm64/Resources/lib/libRlapack.dylib
 lapack_version 3.11.0

─ Python configuration ─────────────────────────────────────────────────────────
 Python is not available

────────────────────────────────────────────────────────────────────────────────
--------------------------------------------------------------------------------
-                               Masked Functions                               -
--------------------------------------------------------------------------------
function `filter` from {package:stats} by package:dplyr
function `lag` from {package:stats} by package:dplyr
function `intersect` from {package:base} by package:dplyr
function `setdiff` from {package:base} by package:dplyr
function `setequal` from {package:base} by package:dplyr
function `union` from {package:base} by package:dplyr
function `plot` from {package:base} by package:graphics
function `body<-` from {package:base} by package:methods
function `kronecker` from {package:base} by package:methods
--------------------------------------------------------------------------------
-                          Used Package and Functions                          -
--------------------------------------------------------------------------------
{package:base} library, mean, summary
{package:dplyr} %>%, group_by, summarise
{package:stats} lm
--------------------------------------------------------------------------------
-                         Program Run Time Information                         -
--------------------------------------------------------------------------------
Start time: 2024-04-08 14:15:26 BST
End time: 2024-04-08 14:15:26 BST
Run time: 0 seconds
--------------------------------------------------------------------------------
-                             Errors and Warnings                              -
--------------------------------------------------------------------------------
Errors:


Warnings:

--------------------------------------------------------------------------------
-                         Messages, Output, and Result                         -
--------------------------------------------------------------------------------
Messages:

Output:


Result:

	Call:
	lm(formula = mpg ~ wt, data = mtcars)

	Residuals:
	    Min      1Q  Median      3Q     Max
	-4.5432 -2.3647 -0.1252  1.4096  6.8727

	Coefficients:
	            Estimate Std. Error t value Pr(>|t|)
	(Intercept)  37.2851     1.8776  19.858  < 2e-16 ***
	wt           -5.3445     0.5591  -9.559 1.29e-10 ***
	---
	Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

	Residual standard error: 3.046 on 30 degrees of freedom
	Multiple R-squared:  0.7528,	Adjusted R-squared:  0.7446
	F-statistic: 91.38 on 1 and 30 DF,  p-value: 1.294e-10

--------------------------------------------------------------------------------
-                               Log Output File                                -
--------------------------------------------------------------------------------
Log name: script.log
Log path: /Users/davidblair/My Drive/Appsilon/Work Projects/examples/logging

```

## `whirl`

Use `whirl::run()` to execute scripts and create logs:

```{r}
#| label: whirl setup
#| include: false
options(whirl.verbosity_level = "minimal")
```

```{r}
#| label: whirl run
whirl::run("logging/script.R", summary_file = NULL)
```

When running scripts the overall status is reported in the console. In this example we are choosing not to create a summary log, since we are only running a single script.

The HTML log is created with the name `script_log.html` and can be viewed below:

```{=html}
<iframe src="script_log.html" width=100% height=600></iframe>
```

:::

### Information in the Log

The `logrx` package offers more detailed output compared to `logr`, as outlined in the following table. A key distinction is that `logrx` automatically logs code output. In `logr`, the user has control over what does or does not get logged. The user can specify which variables, objects, or datasets to send to the log. If the 'autolog' feature is activated, output from dplyr and sassy functions are logged automatically.

|                      | logr | logrx | whirl |
|----------------------|------|-------|-------|
| Log Path             | ✔    | ✔     | ✔     |
| Working Directory    | ✔    | ✔     |       |
| User Name            | ✔    | ✔     |       |
| R Version            | ✔    | ✔     | ✔     |
| Machine              | ✔    | ✔     | ✔     |
| Operating System     | ✔    | ✔     | ✔     |
| Base Packages        | ✔    | ✔     |       |
| Other Packages       | ✔    | ✔     | ✔     |
| Log Start Time       | ✔    | ✔     |       |
| Log End Time.        | ✔    | ✔     | ✔     |
| Elapsed Time         | ✔    | ✔     |       |
| Warnings and Errors  | ✔    | ✔     | ✔     |
| File Hash            |      | ✔     |       |
| Package Dependencies |      | ✔     | ✔     |
| External Software    |      | ✔     | (✔)*  |
| Masked Functions     |      | ✔     |       |
| Renv status          |      |       | ✔     |
| Summary status       |      |       | ✔     |
| Input/output files   |      |       | ✔     |

* Only selected external software, such as Pandoc, Quarto, and Python versions.

### Linting

The `logrx` tool enables the inclusion of `lintR` output within log messages (line 146 - 165 below). To utilize this feature, configure the `log.rx.lint` global option with a selection of lint rules. A comprehensive list of available rules can be accessed [here](https://lintr.r-lib.org/reference/linters.html).

```{r, eval=FALSE}
options(log.rx.lint = list(
  lintr::line_length_linter(20)
))

library(logrx)
axecute("script.R")
```



```{t}
#| class: output-block
--------------------------------------------------------------------------------
-                                logrx Metadata                                -
--------------------------------------------------------------------------------
This log was generated using logrx 0.3.0
logrx package version: 0.3.0
logrx build: CRAN (R 4.3.1)
logrx link to repository: https://github.com/pharmaverse/logrx
--------------------------------------------------------------------------------
-                          User and File Information                           -
--------------------------------------------------------------------------------
User: davidblair
File Name: script.R
File Path: /Users/davidblair/My Drive/Appsilon/Work Projects/examples/logging
File HashSum: 76f9dd029df25ba4f65de72631fac248fa30d7d5
--------------------------------------------------------------------------------
-                             Session Information                              -
--------------------------------------------------------------------------------
─ Session info ─────────────────────────────────────────────────────────────────
 setting  value
 version  R version 4.3.3 (2024-02-29)
 os       macOS Sonoma 14.4.1
 system   aarch64, darwin20
 ui       X11
 language (EN)
 collate  en_US.UTF-8
 ctype    en_US.UTF-8
 tz       Europe/London
 date     2024-04-08
 pandoc   NA

─ Packages ─────────────────────────────────────────────────────────────────────
 ! package      * version date (UTC) lib source
   backports      1.4.1   2021-12-13 [1] CRAN (R 4.3.0)
   cachem         1.0.8   2023-05-01 [1] CRAN (R 4.3.0)
   callr          3.7.6   2024-03-25 [1] CRAN (R 4.3.1)
   cli            3.6.2   2023-12-11 [1] CRAN (R 4.3.1)
   common         1.1.2   2024-03-05 [1] CRAN (R 4.3.1)
   crayon         1.5.2   2022-09-29 [1] CRAN (R 4.3.0)
   curl           5.2.1   2024-03-01 [1] CRAN (R 4.3.1)
   cyclocomp      1.1.1   2023-08-30 [1] CRAN (R 4.3.0)
   desc           1.4.3   2023-12-10 [1] CRAN (R 4.3.1)
   devtools       2.4.5   2022-10-11 [1] CRAN (R 4.3.0)
   digest         0.6.35  2024-03-11 [1] CRAN (R 4.3.1)
   dplyr        * 1.1.4   2023-11-17 [1] CRAN (R 4.3.1)
   ellipsis       0.3.2   2021-04-29 [1] CRAN (R 4.3.0)
   fansi          1.0.6   2023-12-08 [1] CRAN (R 4.3.1)
   fastmap        1.1.1   2023-02-24 [1] CRAN (R 4.3.0)
   fs             1.6.3   2023-07-20 [1] CRAN (R 4.3.0)
   generics       0.1.3   2022-07-05 [1] CRAN (R 4.3.0)
   glue           1.7.0   2024-01-09 [1] CRAN (R 4.3.1)
   htmltools      0.5.8.1 2024-04-04 [1] CRAN (R 4.3.1)
   htmlwidgets    1.6.4   2023-12-06 [1] CRAN (R 4.3.1)
   httpuv         1.6.15  2024-03-26 [1] CRAN (R 4.3.1)
   later          1.3.2   2023-12-06 [1] CRAN (R 4.3.1)
   lazyeval       0.2.2   2019-03-15 [1] CRAN (R 4.3.0)
   lifecycle      1.0.4   2023-11-07 [1] CRAN (R 4.3.1)
   lintr          3.1.2   2024-03-25 [1] CRAN (R 4.3.1)
 V logr         * 1.3.7   2024-02-24 [1] CRAN (R 4.3.1) (on disk 1.3.6)
   logrx        * 0.3.0   2023-10-17 [1] CRAN (R 4.3.1)
   magrittr       2.0.3   2022-03-30 [1] CRAN (R 4.3.0)
   memoise        2.0.1   2021-11-26 [1] CRAN (R 4.3.0)
   mime           0.12    2021-09-28 [1] CRAN (R 4.3.0)
   miniUI         0.1.1.1 2018-05-18 [1] CRAN (R 4.3.0)
   pillar         1.9.0   2023-03-22 [1] CRAN (R 4.3.0)
   pkgbuild       1.4.4   2024-03-17 [1] CRAN (R 4.3.1)
   pkgconfig      2.0.3   2019-09-22 [1] CRAN (R 4.3.0)
   pkgload        1.3.4   2024-01-16 [1] CRAN (R 4.3.1)
   processx       3.8.4   2024-03-16 [1] CRAN (R 4.3.1)
   profvis        0.3.8   2023-05-02 [1] CRAN (R 4.3.0)
   promises       1.2.1   2023-08-10 [1] CRAN (R 4.3.0)
   ps             1.7.6   2024-01-18 [1] CRAN (R 4.3.1)
   purrr          1.0.2   2023-08-10 [1] CRAN (R 4.3.0)
   R6             2.5.1   2021-08-19 [1] CRAN (R 4.3.0)
   Rcpp           1.0.12  2024-01-09 [1] CRAN (R 4.3.1)
   remotes        2.4.2.1 2023-07-18 [1] CRAN (R 4.3.0)
   rex            1.2.1   2021-11-26 [1] CRAN (R 4.3.0)
   rlang          1.1.3   2024-01-10 [1] CRAN (R 4.3.1)
   rstudioapi     0.15.0  2023-07-07 [1] CRAN (R 4.3.0)
   sessioninfo    1.2.2   2021-12-06 [1] CRAN (R 4.3.0)
   shiny          1.8.1.1 2024-04-02 [1] CRAN (R 4.3.1)
   stringi        1.8.3   2023-12-11 [1] CRAN (R 4.3.1)
   stringr        1.5.1   2023-11-14 [1] CRAN (R 4.3.1)
   tibble         3.2.1   2023-03-20 [1] CRAN (R 4.3.0)
   tidyr          1.3.1   2024-01-24 [1] CRAN (R 4.3.1)
   tidyselect     1.2.1   2024-03-11 [1] CRAN (R 4.3.1)
   urlchecker     1.0.1   2021-11-30 [1] CRAN (R 4.3.0)
   usethis        2.2.3   2024-02-19 [1] CRAN (R 4.3.1)
   utf8           1.2.4   2023-10-22 [1] CRAN (R 4.3.1)
   vctrs          0.6.5   2023-12-01 [1] CRAN (R 4.3.1)
   waiter         0.2.5   2022-01-03 [1] CRAN (R 4.3.0)
   withr          3.0.0   2024-01-16 [1] CRAN (R 4.3.1)
   xml2           1.3.6   2023-12-04 [1] CRAN (R 4.3.1)
   xmlparsedata   1.0.5   2021-03-06 [1] CRAN (R 4.3.0)
   xtable         1.8-4   2019-04-21 [1] CRAN (R 4.3.0)

 [1] /Library/Frameworks/R.framework/Versions/4.3-arm64/Resources/library

 V ── Loaded and on-disk version mismatch.

─ External software ────────────────────────────────────────────────────────────
 setting        value
 cairo
 cairoFT
 pango
 png
 jpeg
 tiff
 tcl
 curl           8.4.0
 zlib           1.2.12
 bzlib          1.0.8, 13-Jul-2019
 xz             5.4.4
 PCRE           10.42 2022-12-11
 ICU            74.1
 TRE            TRE 0.8.0 R_fixes (BSD)
 iconv          Apple or GNU libiconv 1.11
 readline       5.2

	  BLAS           /Library/Frameworks/R.framework/Versions/4.3-arm64/Resources/lib/libRblas.0.dylib

	  lapack         /Library/Frameworks/R.framework/Versions/4.3-arm64/Resources/lib/libRlapack.dylib
 lapack_version 3.11.0

─ Python configuration ─────────────────────────────────────────────────────────
 Python is not available

────────────────────────────────────────────────────────────────────────────────
--------------------------------------------------------------------------------
-                               Masked Functions                               -
--------------------------------------------------------------------------------
function `filter` from {package:stats} by package:dplyr
function `lag` from {package:stats} by package:dplyr
function `intersect` from {package:base} by package:dplyr
function `setdiff` from {package:base} by package:dplyr
function `setequal` from {package:base} by package:dplyr
function `union` from {package:base} by package:dplyr
function `plot` from {package:base} by package:graphics
function `body<-` from {package:base} by package:methods
function `kronecker` from {package:base} by package:methods
--------------------------------------------------------------------------------
-                          Used Package and Functions                          -
--------------------------------------------------------------------------------
{package:base} library, mean, summary
{package:dplyr} %>%, group_by, summarise
{package:stats} lm
--------------------------------------------------------------------------------
-                                Linter Results                                -
--------------------------------------------------------------------------------
Line 1 [line_length_linter] Lines should not be more than 20 characters. This
line is 22 characters.

Line 5 [line_length_linter] Lines should not be more than 20 characters. This
line is 25 characters.

Line 7 [line_length_linter] Lines should not be more than 20 characters. This
line is 25 characters.

Line 9 [line_length_linter] Lines should not be more than 20 characters. This
line is 50 characters.

Line 11 [line_length_linter] Lines should not be more than 20 characters.
This line is 36 characters.

Line 14 [line_length_linter] Lines should not be more than 20 characters.
This line is 27 characters.
--------------------------------------------------------------------------------
-                         Program Run Time Information                         -
--------------------------------------------------------------------------------
Start time: 2024-04-08 15:26:12 BST
End time: 2024-04-08 15:26:12 BST
Run time: 0 seconds
--------------------------------------------------------------------------------
-                             Errors and Warnings                              -
--------------------------------------------------------------------------------
Errors:


Warnings:

--------------------------------------------------------------------------------
-                         Messages, Output, and Result                         -
--------------------------------------------------------------------------------
Messages:

Output:


Result:

	Call:
	lm(formula = mpg ~ wt, data = mtcars)

	Residuals:
	    Min      1Q  Median      3Q     Max
	-4.5432 -2.3647 -0.1252  1.4096  6.8727

	Coefficients:
	            Estimate Std. Error t value Pr(>|t|)
	(Intercept)  37.2851     1.8776  19.858  < 2e-16 ***
	wt           -5.3445     0.5591  -9.559 1.29e-10 ***
	---
	Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

	Residual standard error: 3.046 on 30 degrees of freedom
	Multiple R-squared:  0.7528,	Adjusted R-squared:  0.7446
	F-statistic: 91.38 on 1 and 30 DF,  p-value: 1.294e-10

--------------------------------------------------------------------------------
-                               Log Output File                                -
--------------------------------------------------------------------------------
Log name: script.log
Log path: /Users/davidblair/My Drive/Appsilon/Work Projects/examples/logging
```

### Global Options

We've previously discussed the global options in the `logrx` package that enable the inclusion of `lintr` output in the log. There are several key differences between the global options in `logrx` and `whirl`, and those in `logr`. The options in `logr` primarily focus on managing script output, whereas those in `logrx` and `whirl` influence the scope of information included, regardless of the script output.

::: {.panel-tabset}

## `logr`

1. `logr.autolog`: Enables automatic logging for various `dplyr` and `tidyr` functions.
2. `logr.compact`: Removes any blanks spaces between log entries.
3. `logr.traceback`: Includes the full traceback for errors.
4. `logr.warnings`: Retrieves warnings encountered during log file generation, used with `getOption`.
5. `logr.on`: Controls whether the `logr` log is active.
6. `logr.notes`: Includes notes within the log.

## `logrx`

1. `log.rx`: A blank log environment for storing log elements during execution.
2. `log.rx.exec.env`: The execution environment for program code.
3. `log.rx.lint`: Contains `lintr` functions/rules for script analysis.
4. `log.rx.approved`: Specifies the location of approved functions.

## `whirl`

The most important options to configure `whirl` is listed below:

1. `whirl.verbosity_level`: Controls the level of information displayed in the console.
1. `whirl.n_workers`: How many parallel sessions can `whirl` use when running multiple 
scripts simultanously.
1. `whirl.track_files`: Should reading, writing, and deleting files be tracked?
1. `whirl.check_renv`: Should the R environments compliance with your `renv` lockfile be
be checked?

Most are also available as direct arguments in `whirl::run()`. For the full list 
and more details see the
[options](https://novonordisk-opensource.github.io/whirl/reference/options.html) reference.

:::

### Approved Functions

`logrx` features the ability to designate approved functions, highlighting any functions in the script not on this list within the log file. This enhances traceability of package and function use within a validated environment.

To utilize this feature, begin by creating a list of approved functions:
```{r}
approved_pkgs <- list(base = "mean",
                      dplyr = "All")
```

Next, build the `approved_pkgs` list and save its contents to a file:
```{r}
library(logrx)
build_approved(approved_pkgs, file = "approved.rds")
```

Finally, assign the file to the global option `log.rx.approved`. This ensures the analysis is incorporated into the log (refer to lines 140 - 145 below):
```{r}
options(log.rv.approved = "approved.rds")
```



```{r, eval=FALSE}
axecute("script.R")
```



```{t}
#| class: output-block
--------------------------------------------------------------------------------
-                                logrx Metadata                                -
--------------------------------------------------------------------------------
This log was generated using logrx 0.3.0
logrx package version: 0.3.0
logrx build: CRAN (R 4.3.1)
logrx link to repository: https://github.com/pharmaverse/logrx
--------------------------------------------------------------------------------
-                          User and File Information                           -
--------------------------------------------------------------------------------
User: davidblair
File Name: script.R
File Path: /Users/davidblair/My Drive/Appsilon/Work Projects/examples/logging
File HashSum: 76f9dd029df25ba4f65de72631fac248fa30d7d5
--------------------------------------------------------------------------------
-                             Session Information                              -
--------------------------------------------------------------------------------
─ Session info ─────────────────────────────────────────────────────────────────
 setting  value
 version  R version 4.3.3 (2024-02-29)
 os       macOS Sonoma 14.4.1
 system   aarch64, darwin20
 ui       X11
 language (EN)
 collate  en_US.UTF-8
 ctype    en_US.UTF-8
 tz       Europe/London
 date     2024-04-08
 pandoc   NA

─ Packages ─────────────────────────────────────────────────────────────────────
 ! package      * version date (UTC) lib source
   backports      1.4.1   2021-12-13 [1] CRAN (R 4.3.0)
   cachem         1.0.8   2023-05-01 [1] CRAN (R 4.3.0)
   callr          3.7.6   2024-03-25 [1] CRAN (R 4.3.1)
   cli            3.6.2   2023-12-11 [1] CRAN (R 4.3.1)
   common         1.1.2   2024-03-05 [1] CRAN (R 4.3.1)
   crayon         1.5.2   2022-09-29 [1] CRAN (R 4.3.0)
   curl           5.2.1   2024-03-01 [1] CRAN (R 4.3.1)
   cyclocomp      1.1.1   2023-08-30 [1] CRAN (R 4.3.0)
   desc           1.4.3   2023-12-10 [1] CRAN (R 4.3.1)
   devtools       2.4.5   2022-10-11 [1] CRAN (R 4.3.0)
   digest         0.6.35  2024-03-11 [1] CRAN (R 4.3.1)
   dplyr        * 1.1.4   2023-11-17 [1] CRAN (R 4.3.1)
   ellipsis       0.3.2   2021-04-29 [1] CRAN (R 4.3.0)
   fansi          1.0.6   2023-12-08 [1] CRAN (R 4.3.1)
   fastmap        1.1.1   2023-02-24 [1] CRAN (R 4.3.0)
   fs             1.6.3   2023-07-20 [1] CRAN (R 4.3.0)
   generics       0.1.3   2022-07-05 [1] CRAN (R 4.3.0)
   glue           1.7.0   2024-01-09 [1] CRAN (R 4.3.1)
   htmltools      0.5.8.1 2024-04-04 [1] CRAN (R 4.3.1)
   htmlwidgets    1.6.4   2023-12-06 [1] CRAN (R 4.3.1)
   httpuv         1.6.15  2024-03-26 [1] CRAN (R 4.3.1)
   later          1.3.2   2023-12-06 [1] CRAN (R 4.3.1)
   lazyeval       0.2.2   2019-03-15 [1] CRAN (R 4.3.0)
   lifecycle      1.0.4   2023-11-07 [1] CRAN (R 4.3.1)
   lintr          3.1.2   2024-03-25 [1] CRAN (R 4.3.1)
 V logr         * 1.3.7   2024-02-24 [1] CRAN (R 4.3.1) (on disk 1.3.6)
   logrx        * 0.3.0   2023-10-17 [1] CRAN (R 4.3.1)
   magrittr       2.0.3   2022-03-30 [1] CRAN (R 4.3.0)
   memoise        2.0.1   2021-11-26 [1] CRAN (R 4.3.0)
   mime           0.12    2021-09-28 [1] CRAN (R 4.3.0)
   miniUI         0.1.1.1 2018-05-18 [1] CRAN (R 4.3.0)
   pillar         1.9.0   2023-03-22 [1] CRAN (R 4.3.0)
   pkgbuild       1.4.4   2024-03-17 [1] CRAN (R 4.3.1)
   pkgconfig      2.0.3   2019-09-22 [1] CRAN (R 4.3.0)
   pkgload        1.3.4   2024-01-16 [1] CRAN (R 4.3.1)
   processx       3.8.4   2024-03-16 [1] CRAN (R 4.3.1)
   profvis        0.3.8   2023-05-02 [1] CRAN (R 4.3.0)
   promises       1.2.1   2023-08-10 [1] CRAN (R 4.3.0)
   ps             1.7.6   2024-01-18 [1] CRAN (R 4.3.1)
   purrr          1.0.2   2023-08-10 [1] CRAN (R 4.3.0)
   R6             2.5.1   2021-08-19 [1] CRAN (R 4.3.0)
   Rcpp           1.0.12  2024-01-09 [1] CRAN (R 4.3.1)
   remotes        2.4.2.1 2023-07-18 [1] CRAN (R 4.3.0)
   rex            1.2.1   2021-11-26 [1] CRAN (R 4.3.0)
   rlang          1.1.3   2024-01-10 [1] CRAN (R 4.3.1)
   rstudioapi     0.15.0  2023-07-07 [1] CRAN (R 4.3.0)
   sessioninfo    1.2.2   2021-12-06 [1] CRAN (R 4.3.0)
   shiny          1.8.1.1 2024-04-02 [1] CRAN (R 4.3.1)
   stringi        1.8.3   2023-12-11 [1] CRAN (R 4.3.1)
   stringr        1.5.1   2023-11-14 [1] CRAN (R 4.3.1)
   tibble         3.2.1   2023-03-20 [1] CRAN (R 4.3.0)
   tidyr          1.3.1   2024-01-24 [1] CRAN (R 4.3.1)
   tidyselect     1.2.1   2024-03-11 [1] CRAN (R 4.3.1)
   urlchecker     1.0.1   2021-11-30 [1] CRAN (R 4.3.0)
   usethis        2.2.3   2024-02-19 [1] CRAN (R 4.3.1)
   utf8           1.2.4   2023-10-22 [1] CRAN (R 4.3.1)
   vctrs          0.6.5   2023-12-01 [1] CRAN (R 4.3.1)
   waiter         0.2.5   2022-01-03 [1] CRAN (R 4.3.0)
   withr          3.0.0   2024-01-16 [1] CRAN (R 4.3.1)
   xml2           1.3.6   2023-12-04 [1] CRAN (R 4.3.1)
   xmlparsedata   1.0.5   2021-03-06 [1] CRAN (R 4.3.0)
   xtable         1.8-4   2019-04-21 [1] CRAN (R 4.3.0)

 [1] /Library/Frameworks/R.framework/Versions/4.3-arm64/Resources/library

 V ── Loaded and on-disk version mismatch.

─ External software ────────────────────────────────────────────────────────────
 setting        value
 cairo
 cairoFT
 pango
 png
 jpeg
 tiff
 tcl
 curl           8.4.0
 zlib           1.2.12
 bzlib          1.0.8, 13-Jul-2019
 xz             5.4.4
 PCRE           10.42 2022-12-11
 ICU            74.1
 TRE            TRE 0.8.0 R_fixes (BSD)
 iconv          Apple or GNU libiconv 1.11
 readline       5.2

	  BLAS           /Library/Frameworks/R.framework/Versions/4.3-arm64/Resources/lib/libRblas.0.dylib

	  lapack         /Library/Frameworks/R.framework/Versions/4.3-arm64/Resources/lib/libRlapack.dylib
 lapack_version 3.11.0

─ Python configuration ─────────────────────────────────────────────────────────
 Python is not available

────────────────────────────────────────────────────────────────────────────────
--------------------------------------------------------------------------------
-                               Masked Functions                               -
--------------------------------------------------------------------------------
function `filter` from {package:stats} by package:dplyr
function `lag` from {package:stats} by package:dplyr
function `intersect` from {package:base} by package:dplyr
function `setdiff` from {package:base} by package:dplyr
function `setequal` from {package:base} by package:dplyr
function `union` from {package:base} by package:dplyr
function `plot` from {package:base} by package:graphics
function `body<-` from {package:base} by package:methods
function `kronecker` from {package:base} by package:methods
--------------------------------------------------------------------------------
-                          Used Package and Functions                          -
--------------------------------------------------------------------------------
{package:base} library, mean, summary
{package:dplyr} %>%, group_by, summarise
{package:stats} lm
--------------------------------------------------------------------------------
-                       Unapproved Package and Functions                       -
--------------------------------------------------------------------------------
{package:base} library, summary
{package:stats} lm
--------------------------------------------------------------------------------
-                                Linter Results                                -
--------------------------------------------------------------------------------
Line 1 [line_length_linter] Lines should not be more than 20 characters. This
line is 22 characters.

Line 5 [line_length_linter] Lines should not be more than 20 characters. This
line is 25 characters.

Line 7 [line_length_linter] Lines should not be more than 20 characters. This
line is 25 characters.

Line 9 [line_length_linter] Lines should not be more than 20 characters. This
line is 50 characters.

Line 11 [line_length_linter] Lines should not be more than 20 characters.
This line is 36 characters.

Line 14 [line_length_linter] Lines should not be more than 20 characters.
This line is 27 characters.
--------------------------------------------------------------------------------
-                         Program Run Time Information                         -
--------------------------------------------------------------------------------
Start time: 2024-04-08 16:03:28 BST
End time: 2024-04-08 16:03:28 BST
Run time: 0 seconds
--------------------------------------------------------------------------------
-                             Errors and Warnings                              -
--------------------------------------------------------------------------------
Errors:


Warnings:

--------------------------------------------------------------------------------
-                         Messages, Output, and Result                         -
--------------------------------------------------------------------------------
Messages:

Output:


Result:

	Call:
	lm(formula = mpg ~ wt, data = mtcars)

	Residuals:
	    Min      1Q  Median      3Q     Max
	-4.5432 -2.3647 -0.1252  1.4096  6.8727

	Coefficients:
	            Estimate Std. Error t value Pr(>|t|)
	(Intercept)  37.2851     1.8776  19.858  < 2e-16 ***
	wt           -5.3445     0.5591  -9.559 1.29e-10 ***
	---
	Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

	Residual standard error: 3.046 on 30 degrees of freedom
	Multiple R-squared:  0.7528,	Adjusted R-squared:  0.7446
	F-statistic: 91.38 on 1 and 30 DF,  p-value: 1.294e-10

--------------------------------------------------------------------------------
-                               Log Output File                                -
--------------------------------------------------------------------------------
Log name: script.log
Log path: /Users/davidblair/My Drive/Appsilon/Work Projects/examples/logging
```

### Run multiple scripts

A key functionality `whirl` is to support the execution of analysis projects. 
It is possible to define batches of scripts to run simultanously, and the order 
in which the different steps should be executed.

It is all controlled by a yaml configuration file (use `whirl::use_whirl()` to get started) such as the one below:

`_whirl.yaml`:
```yaml
steps:
  - name: "Initial step"
    paths:
      - "setup.R"
  - name: "First batch of programs"
    paths:
      - "prgA.R"
      - "prgB.R"
  - name: "Run all programs with 2 prefix"
    paths:
      - "2_*.R"
```

Here three steps are defined, where the first step is to run `setup.R`. After this we want to run `prgA.R` and `prgB.R` simultanously, and after these two have finished we run all R scripts starting with `2_` (imagine they depend on previous output from e.g. `prgA.R`).

To run all steps you just call `run()` with the configuration file as input:

```{r}
#| eval: false
whirl::run("_whirl.yaml")
```

In this scenario it becomes relevant to check the summary log as well (defaults to `summary.html`), which have the execution status of each script, and a link to the individual logs.

### Summary

The decision between using `logr` and `logrx` hinges on specific needs. `logr` offers greater control over the output, at the expense of detailed environmental context. Conversely, `logrx` provides more comprehensive details and quicker log generation, along with better support for command line execution, but offers less control over the output.
