---
title: "DM"
order: 2
---

```{r setup script, include=FALSE, purl=FALSE}
invisible_hook_purl <- function(before, options, ...) {
  knitr::hook_purl(before, options, ...)
  NULL
}
knitr::knit_hooks$set(purl = invisible_hook_purl)
source("functions/print_df.R")
```

# Introduction

This article describes how to create a demographics(DM) domain using the {sdtm.oak} package.

Before reading this article, it is recommended that users review the "Creating an Interventions Domain" article[here]https://pharmaverse.github.io/sdtm.oak/articles/interventions_domain.html, which provides a detailed explanation of various concepts in {sdtm.oak}, such as `oak_id_vars`, `condition_add`, etc. It also offers guidance on which mapping algorithms or functions to use for different mappings and provides a more detailed explanation of how these mapping algorithms or functions work.

In this article, we will dive directly into programming and provide further explanation only where it is required.

# Programming workflow

* [Read in data](#readdata)
* [Create oak_id_vars](#oakidvars)
* [Read in CT](#readct)
* [Create reference dates configuration file](#refdates)
* [Map Topic Variable](#maptopic)
* [Map Rest of the Variables](#maprest)
* [Map Reference Date Variables](#mapvars)
* [Create SDTM derived variables](#derivedvars)
* [Add Labels and Attributes](#attributes)

## Read in data {#readdata}

Read all the raw datasets into the environment. In this example, the raw datasets needed are `ec_raw`, `ds_raw` and `dm_raw`. Users can read them from the `{pharmaverseraw}` package using the below code:

```{r setup, message=FALSE, warning=FALSE, results='hold'}
library(sdtm.oak)
library(pharmaverseraw)
library(dplyr)

dm_raw <- pharmaverseraw::dm_raw
ds_raw <- pharmaverseraw::ds_raw
ec_raw <- pharmaverseraw::ec_raw
```

Demographics Raw dataset. The mock up of the CRF can be viewed [here](https://github.com/pharmaverse/pharmaverseraw/blob/main/vignettes/articles/aCRFs/Demographics_aCRF.pdf):
```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(dm_raw, n = 5)
```


Disposition Raw dataset. The mock up of the CRF can be viewed [here](https://github.com/pharmaverse/pharmaverseraw/blob/main/vignettes/articles/aCRFs/Exposure_as_collected_aCRF.pdf)
```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(ds_raw, n = 5)
```


Study Drug Administration Raw dataset. The mock up of the CRF can be viewed [here](https://github.com/pharmaverse/pharmaverseraw/blob/main/vignettes/articles/aCRFs/Subject_Disposition_aCRF.pdf)
```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(ec_raw, n = 5)
```

## Create oak_id_vars {#oakidvars}

```{r}
dm_raw <- dm_raw %>%
  generate_oak_id_vars(
    pat_var = "PATNUM",
    raw_src = "dm_raw"
  )

ds_raw <- ds_raw %>%
  generate_oak_id_vars(
    pat_var = "PATNUM",
    raw_src = "ds_raw"
  )

ec_raw <- ec_raw %>%
  generate_oak_id_vars(
    pat_var = "PATNUM",
    raw_src = "ec_raw"
  )
```

For example, Demographics Raw dataset with `oak_id_vars`
```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(dm_raw, n = 10)
```

## Read in CT {#readct}

Controlled Terminology is part of the SDTM specification and it is prepared by the user.  In this example, the study controlled terminology name is `sdtm_ct.csv`. Users can read it from the package using the below code:
```{r, echo = TRUE}
# TODO: update file path when the file is moved to the correct folder
study_ct <- read.csv("metadata/sdtm_ct.csv")
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(study_ct, n = 10)
```

## Create reference dates configuration file {#refdates}

Create reference date configuration file, a data frame which has the details of the variables to be used for the calculation of reference dates. The data frame should have columns listed below:
- **`raw_dataset_name`**: Name of the raw dataset.  
- **`date_var`**: Date variable name from the raw dataset.  
- **`time_var`**: Time variable name from the raw dataset.  
- **`dformat`**: Format of the date collected in raw data.  
- **`tformat`**: Format of the time collected in raw data.  
- **`sdtm_var_name`**: Reference date variable name in DM domain where the raw variable is used.  

```{r}
ref_date_conf_df <- tibble::tribble(
  ~raw_dataset_name, ~date_var,     ~time_var,      ~dformat,      ~tformat, ~sdtm_var_name,
  "ec_raw",       "IT.ECSTDAT", NA_character_, "dd-mmm-yyyy", NA_character_,     "RFXSTDTC",
  "ec_raw",       "IT.ECENDAT", NA_character_, "dd-mmm-yyyy", NA_character_,     "RFXENDTC",
  "ec_raw",       "IT.ECSTDAT", NA_character_, "dd-mmm-yyyy", NA_character_,      "RFSTDTC",
  "ec_raw",       "IT.ECENDAT", NA_character_, "dd-mmm-yyyy", NA_character_,      "RFENDTC",
  "dm_raw",            "IC_DT", NA_character_,  "mm/dd/yyyy", NA_character_,      "RFICDTC",
  "ds_raw",          "DSDTCOL",     "DSTMCOL",  "mm-dd-yyyy",         "H:M",     "RFPENDTC",
  "ds_raw",          "DEATHDT", NA_character_,  "mm/dd/yyyy", NA_character_,       "DTHDTC"
)
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(ref_date_conf_df)
```

## Map Topic Variable {#maptopic}

In DM domain SUBJID is the topic variable and it can be mapped from PATNUM using a simple mutate statement.

```{r}
dm <- dm_raw %>%
  mutate(
    SUBJID = substr(PATNUM, 5, 8)
  ) %>%
  select(oak_id, raw_source, patient_number, SUBJID)
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(dm, n = 10)
```

## Map Rest of the Variables {#maprest}

Map rest of the variables in DM domain using either `assign_no_ct` or `assign_ct` depending
on if the variable has controlled terminologies associated.

```{r}
dm <- dm %>%
  # Map AGE using assign_no_ct
  assign_no_ct(
    raw_dat = dm_raw,
    raw_var = "IT.AGE",
    tgt_var = "AGE",
    id_vars = oak_id_vars()
  ) %>%
  # Map AGEU using hardcode_ct
  hardcode_ct(
    raw_dat = dm_raw,
    raw_var = "IT.AGE",
    tgt_var = "AGEU",
    tgt_val = "Year",
    ct_spec = study_ct,
    ct_clst = "C66781",
    id_vars = oak_id_vars()
  ) %>%
  # Map SEX using assign_ct
  assign_ct(
    raw_dat = dm_raw,
    raw_var = "IT.SEX",
    tgt_var = "SEX",
    ct_spec = study_ct,
    ct_clst = "C66731",
    id_vars = oak_id_vars()
  ) %>%
  # Map ETHNIC using assign_ct
  assign_ct(
    raw_dat = dm_raw,
    raw_var = "IT.ETHNIC",
    tgt_var = "ETHNIC",
    ct_spec = study_ct,
    ct_clst = "C66790",
    id_vars = oak_id_vars()
  ) %>%
  # Map RACE using assign_ct
  assign_ct(
    raw_dat = dm_raw,
    raw_var = "IT.RACE",
    tgt_var = "RACE",
    ct_spec = study_ct,
    ct_clst = "C74457",
    id_vars = oak_id_vars()
  ) %>%
  # Map ARM using assign_ct
  assign_ct(
    raw_dat = dm_raw,
    raw_var = "PLANNED_ARM",
    tgt_var = "ARM",
    ct_spec = study_ct,
    ct_clst = "ARM",
    id_vars = oak_id_vars()
  ) %>%
  # Map ARMCD using assign_no_ct
  assign_no_ct(
    raw_dat = dm_raw,
    raw_var = "PLANNED_ARMCD",
    tgt_var = "ARMCD",
    id_vars = oak_id_vars()
  ) %>%
  # Map ACTARM using assign_ct
  assign_ct(
    raw_dat = dm_raw,
    raw_var = "ACTUAL_ARM",
    tgt_var = "ACTARM",
    ct_spec = study_ct,
    ct_clst = "ARM",
    id_vars = oak_id_vars()
  ) %>%
  # Map ACTARMCD using assign_no_ct
  assign_no_ct(
    raw_dat = dm_raw,
    raw_var = "ACTUAL_ARMCD",
    tgt_var = "ACTARMCD",
    id_vars = oak_id_vars()
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(dm, n = 10)
```

## Map Reference Date Variables {#mapvars}

Use `oak_cal_ref_dates` to calculate reference dates variables in ISO8601 format.
The function takes the raw variable names from reference date configuration file,
and calculated the minimum or maximum dates based upon the `min_max` parameter.

Variable `RFSTDTC` is the reference Start Date/time for the subject in ISO 8601 character format. Usually equivalent to date/time when subject was first exposed to study treatment. So as specified in the reference date configuration file, we need to calculate the minumum date of  the `IT.ECSTDAT` for each subject from the `ec_raw` dataset. Therefore, in `min_max` parameter, "min" is selected for the calculation.

```{r eval=TRUE}
dm <- dm %>%
  # Derive RFSTDTC using oak_cal_ref_dates
  oak_cal_ref_dates(
    ds_in = .,
    der_var = "RFSTDTC",
    min_max = "min",
    ref_date_config_df = ref_date_conf_df,
    raw_source = list(
      ec_raw = ec_raw,
      ds_raw = ds_raw,
      dm_raw = dm_raw
    )
  )
```

Variable `RFENDTC` is the Reference end date/time for the subject in ISO 8601 character format. Usually equivalent to the date/time when subject was determined to have ended the trial, and often equivalent to date/time of last exposure to study treatment. As  specified in the reference date configuration file, we need to calculate the  maximum date of the `IT.ECENDAT` for each subject from the `ec_raw` dataset. 
Therefore, in `min_max` parameter, "max" is selected for the calculation. 

```{r}
dm <- dm %>%
  # Derive RFENDTC using oak_cal_ref_dates
  oak_cal_ref_dates(
    ds_in = .,
    der_var = "RFENDTC",
    min_max = "max",
    ref_date_config_df = ref_date_conf_df,
    raw_source = list(
      ec_raw = ec_raw,
      ds_raw = ds_raw,
      dm_raw = dm_raw
    )
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(dm, n = 10)
```

The same derivation logic is applicable to other reference date/time variables.

```{r}
dm <- dm %>%
  # Derive RFXSTDTC using oak_cal_ref_dates
  oak_cal_ref_dates(
    ds_in = .,
    der_var = "RFXSTDTC",
    min_max = "min",
    ref_date_config_df = ref_date_conf_df,
    raw_source = list(
      ec_raw = ec_raw,
      ds_raw = ds_raw,
      dm_raw = dm_raw
    )
  ) %>%
  # Derive RFXENDTC using oak_cal_ref_dates
  oak_cal_ref_dates(
    ds_in = .,
    der_var = "RFXENDTC",
    min_max = "max",
    ref_date_config_df = ref_date_conf_df,
    raw_source = list(
      ec_raw = ec_raw,
      ds_raw = ds_raw,
      dm_raw = dm_raw
    )
  ) %>%
  # Derive RFICDTC using oak_cal_ref_dates
  oak_cal_ref_dates(
    ds_in = .,
    der_var = "RFICDTC",
    min_max = "min",
    ref_date_config_df = ref_date_conf_df,
    raw_source = list(
      ec_raw = ec_raw,
      ds_raw = ds_raw,
      dm_raw = dm_raw
    )
  ) %>%
  # Derive RFPENDTC using oak_cal_ref_dates
  oak_cal_ref_dates(
    ds_in = .,
    der_var = "RFPENDTC",
    min_max = "max",
    ref_date_config_df = ref_date_conf_df,
    raw_source = list(
      ec_raw = ec_raw,
      ds_raw = ds_raw,
      dm_raw = dm_raw
    )
  ) %>%
  # Map DTHDTC using oak_cal_ref_dates
  oak_cal_ref_dates(
    ds_in = .,
    der_var = "DTHDTC",
    min_max = "min",
    ref_date_config_df = ref_date_conf_df,
    raw_source = list(
      ec_raw = ec_raw,
      ds_raw = ds_raw,
      dm_raw = dm_raw
    )
  )
```

## Create SDTM derived variables {#derivedvars}

```{r}
dm <- dm %>%
  mutate(
    STUDYID = dm_raw$STUDY,
    DOMAIN = "DM",
    USUBJID = paste0("01-", dm_raw$PATNUM),
    COUNTRY = dm_raw$COUNTRY,
    DTHFL = dplyr::if_else(is.na(DTHDTC), NA_character_, "Y")
  ) %>%
  # Map DMDTC using assign_datetime
  assign_datetime(
    raw_dat = dm_raw,
    raw_var = "COL_DT",
    tgt_var = "DMDTC",
    raw_fmt = c("m/d/y"),
    id_vars = oak_id_vars()
  ) %>%
  # Derive study day
  derive_study_day(
    sdtm_in = .,
    dm_domain = .,
    tgdt = "DMDTC",
    refdt = "RFXSTDTC",
    study_day_var = "DMDY"
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(dm, n = 10)
```

## Add Labels and Attributes {#attributes}

Yet to be developed.
