---
title: "PP"
order: 4
---

```{r setup script, include=FALSE, purl=FALSE}
invisible_hook_purl <- function(before, options, ...) {
  knitr::hook_purl(before, options, ...)
  NULL
}
knitr::knit_hooks$set(purl = invisible_hook_purl)
source("functions/print_df.R")
```

# Introduction

This article shows how to create a Pharmacokinetic Parameters (`PP`) domain
from an ADNCA dataset using the `{aNCA}` package. `PP` is the SDTM domain that
holds non-compartmental analysis (NCA) results — one row per parameter per
subject per time-reference.

`{aNCA}` wraps `{PKNCA}` for the NCA calculations and adds column mapping,
CDISC code translation, flag rules and export helpers so the full pipeline
runs in a single script.

By the end of this example you will have:

- NCA results computed from concentration-time data
- A CDISC-compliant **PP** dataset
- Companion **ADPP** and **ADNCA** datasets

# Programming workflow

* [Load packages and data](#load)
* [Define column mapping](#mapping)
* [Set up filters](#filters)
* [Define NCA settings](#settings)
* [Create and configure the PKNCA data object](#create)
* [Run NCA calculations](#run)
* [Apply flag rules](#flags)
* [Export CDISC datasets (PP, ADPP, ADNCA)](#export)

## Load packages and data {#load}

```{r load, eval=FALSE}
library(aNCA)
library(dplyr)
```

`{aNCA}` ships with `adnca_example`, a 76-row ADNCA dataset with two analytes
(DrugA and Metab-DrugA), two specimen types (SERUM and URINE) and two dosing
periods. Subjects received different dose levels via extravascular and
intravascular routes.

```{r data, eval=FALSE}
adnca_data <- adnca_example
head(adnca_data)
```

## Define column mapping {#mapping}

The mapping tells `{aNCA}` which columns in your dataset correspond to the
expected CDISC variables. Keys are standard ADNCA column names; values are the
actual column names in your data.

```{r mapping, eval=FALSE}
mapping <- list(
  STUDYID  = "STUDYID",
  USUBJID  = "USUBJID",
  DOSEA    = "DOSEA",
  DOSEU    = "DOSEU",
  DOSETRT  = "DOSETRT",
  PARAM    = "PARAM",
  Metabolites = "Metab-DrugA",
  ARRLT    = "ARRLT",
  NRRLT    = "NRRLT",
  AFRLT    = "AFRLT",
  NCAwXRS  = c("NCA1XRS", "NCA2XRS"),
  NFRLT    = "NFRLT",
  PCSPEC   = "PCSPEC",
  ROUTE    = "ROUTE",
  TRTRINT  = "TRTRINT",
  ADOSEDUR = "ADOSEDUR",
  Grouping_Variables = c("TRT01A", "RACE", "SEX"),
  RRLTU    = "RRLTU",
  VOLUME   = "VOLUME",
  VOLUMEU  = "VOLUMEU",
  AVAL     = "AVAL",
  AVALU    = "AVALU",
  ATPTREF  = "ATPTREF"
)
```

A few notes:

- **`Metabolites`**: PARAM value(s) that represent metabolites. Sets the
  `METABFL` flag internally.
- **`NCAwXRS`**: Column(s) with NCA exclusion reason codes. Flagged subjects
  are excluded from calculations.
- **`Grouping_Variables`**: Columns carried through to results and used for
  grouping in descriptive statistics.

## Set up filters {#filters}

Filters restrict the data before analysis. Each filter specifies a column and
the values to keep. Here we keep only the parent drug for serum specimens:

```{r filters, eval=FALSE}
applied_filters <- list(
  list(column = "PARAM",  value = c("DrugA")),
  list(column = "PCSPEC", value = c("SERUM"))
)
```

## Define NCA settings {#settings}

### Partial interval calculations

Interval parameters compute AUC or average concentration over custom time
windows:

```{r int_parameters, eval=FALSE}
int_parameters <- data.frame(
  parameter = c("AUCINT", "CAVGINT"),
  start_auc = c(0, 0),
  end_auc   = c(12, 12)
)
```

### Parameter selections

Select which NCA parameters to compute, grouped by study type:

```{r parameters, eval=FALSE}
parameters_selected_per_study <- list(
  `Single Extravascular` = c(
    "auclast", "cmax", "tmax", "half.life", "aucinf.obs",
    "vz.obs", "cl.obs", "aumclast"
  ),
  `Single IV Bolus` = c(
    "auclast", "cmax", "half.life", "aucinf.obs",
    "vz.obs", "cl.obs"
  )
)
```

### Custom units

Override default units for specific parameters:

```{r units, eval=FALSE}
units_table <- data.frame(
  PPTESTCD = c("CMAX", "TMAX", "AUCLAST"),
  PPSTRESU = c("ng/mL", "h", "ng*h/mL")
)
```

### Flag rules

Quality thresholds for terminal elimination regression. Parameters that violate
a checked rule are flagged and excluded from summary statistics:

```{r flags, eval=FALSE}
flag_rules <- list(
  R2ADJ   = list(is.checked = TRUE,  threshold = 0.7),
  R2      = list(is.checked = FALSE, threshold = 0.7),
  AUCPEO  = list(is.checked = TRUE,  threshold = 20),
  AUCPEP  = list(is.checked = TRUE,  threshold = 20),
  LAMZSPN = list(is.checked = TRUE,  threshold = 2)
)
```

### Slope rules and extra variables

Slope rules allow manual override of the lambda-z regression range. Pass an
empty data frame if no manual adjustments are needed:

```{r slope_extra, eval=FALSE}
slope_rules <- data.frame(
  USUBJID = character(0), ATPTREF = character(0),
  DOSNOA  = character(0), TYPE    = character(0),
  RANGE   = character(0), REASON  = character(0)
)

extra_vars_to_keep <- c(
  mapping$Grouping_Variables, "DOSEA", "ATPTREF", "ROUTE"
)
```

## Create and configure the PKNCA data object {#create}

`PKNCA_create_data_object()` applies the column mapping, filters and duplicate
handling. `PKNCA_update_data_object()` then configures the AUC method,
intervals, parameters and units:

```{r create, eval=FALSE}
pknca_obj <- adnca_data %>%
  PKNCA_create_data_object(
    mapping = mapping,
    applied_filters = applied_filters,
    time_duplicate_rows = NULL
  ) %>%
  PKNCA_update_data_object(
    method = "lin up/log down",
    selected_analytes = "DrugA",
    selected_profile = "DOSE 1",
    selected_pcspec = "SERUM",
    start_impute = "yes",
    exclusion_list = list(),
    hl_adj_rules = slope_rules,
    keep_interval_cols = setdiff(extra_vars_to_keep, c("DOSEA", "ATPTREF", "ROUTE")),
    min_hl_points = 3,
    parameter_selections = parameters_selected_per_study,
    int_parameters = int_parameters,
    custom_units_table = units_table
  )
```

## Run NCA calculations {#run}

```{r run, eval=FALSE}
pknca_res <- pknca_obj %>%
  PKNCA_calculate_nca(
    blq_rule = list(first = "keep", middle = "keep", last = "keep")
  ) %>%
  add_f_to_pknca_results(NULL) %>%
  mutate(PPTESTCD = translate_terms(PPTESTCD, "PKNCA", "PPTESTCD"))
```

- **`blq_rule`**: How to handle BLQ values at different positions in the
  concentration-time profile. Options: `"keep"`, `"0"`, `"loq/2"`.
- **`add_f_to_pknca_results(NULL)`**: Pass an AUC type to compute
  bioavailability, or `NULL` to skip.

## Apply flag rules {#flags}

Flag rules mark parameters that fail quality thresholds. Flagged parameters
get an exclusion reason and are dropped from descriptive statistics:

```{r apply_flags, eval=FALSE}
pknca_res <- pknca_res %>%
  PKNCA_hl_rules_exclusion(
    rules = flag_rules %>%
      purrr::keep(\(x) x$is.checked) %>%
      purrr::map(\(x) x$threshold)
  )
```

## Export CDISC datasets (PP, ADPP, ADNCA) {#export}

`export_cdisc()` produces the PP, ADPP and ADNCA datasets. Flag rules are
formatted into CRITy/CRITyFL/PPSUMFL columns in ADPP:

```{r export, eval=FALSE}
# Format flag rule messages for ADPP criterion columns
flag_operators <- c(
  R2ADJ = " < ", R2 = " < ", AUCPEO = " > ", AUCPEP = " > ", LAMZSPN = " < "
)
checked_flags <- purrr::keep(flag_rules, function(x) x$is.checked)
flag_rule_msgs <- if (length(checked_flags) > 0) {
  vapply(names(checked_flags), function(nm) {
    paste0(nm, flag_operators[nm], checked_flags[[nm]]$threshold)
  }, character(1), USE.NAMES = FALSE)
} else {
  NULL
}

cdisc_datasets <- pknca_res %>%
  export_cdisc(
    grouping_vars = extra_vars_to_keep,
    flag_rules = flag_rule_msgs
  )
```

The result is a named list with three data frames:

- **`cdisc_datasets$pp`** — PP domain: one row per parameter per subject.
- **`cdisc_datasets$adpp`** — ADPP: PP with ADaM-style flags and criterion
  columns (CRITy, CRITyFL, PPSUMFL, PPSUMRSN).
- **`cdisc_datasets$adnca`** — ADNCA: the full analysis dataset with results
  joined back to the concentration-time data.

```{r view_pp, eval=FALSE}
# View the PP domain
head(cdisc_datasets$pp)
```

# Alternative workflows

## Generate the script from a YAML settings file

The aNCA Shiny app exports a settings YAML that captures the full analysis
configuration. You can regenerate the R script from it:

```{r yaml, eval=FALSE}
get_settings_code(
  settings_file_path = "my_settings.yaml",
  data_path = "my_adnca_data.csv",
  output_path = "nca_script.R"
)
```

## Generate from the aNCA Shiny app

```{r app, eval=FALSE}
aNCA::run_app()
```

From the app, configure mapping, filters and NCA settings, then export via the
**Save** button to get the settings YAML, reproducible R script and all outputs.
