---
title: "ADAE"
order: 7
---

```{r setup script, include=FALSE, purl=FALSE}
invisible_hook_purl <- function(before, options, ...) {
  knitr::hook_purl(before, options, ...)
  NULL
}
knitr::knit_hooks$set(purl = invisible_hook_purl)
source("functions/print_df.R")
```

# Introduction

This article provides a step-by-step explanation for creating an ADaM `ADAE` (Adverse Events) dataset using key pharmaverse packages along with tidyverse components.

Here we try to cover the most common AE analysis dataset derivations and some of the most useful functions for these. 

For the purpose of this example, we will use the `ADSL` dataset from `{pharmaverseadam}` and `ae` and `ex` domains from `{pharmaversesdtm}`.

## Programming Flow

* [Load Data and Required pharmaverse Packages](#loaddata)
* [Load Specifications for Metacore](#loadspecs)
* [Select ADSL Variables](#adslvars)
* [Start Building Derivations](#startbuild)
* [Derive Analysis Dates](#anldates)
* [Derive Duration](#anldur)
* [Derive Date of Last Dose](#exdates)
* [Derive Analysis Flags](#anlflags)
* [Derive Occurrence Flags](#occflags)
* [Derive Query Variables](#query)
* [Add ADSL Variables](#adsladd)
* [Apply Metadata to Create an eSub XPT and Perform Associated Checks](#metacore_xportr)

# Load Data and Required pharmaverse Packages {#loaddata}

The first step is to load the required packages and input data.

```{r setup, message=FALSE, warning=FALSE, results='hold'}
library(metacore)
library(metatools)
library(pharmaversesdtm)
library(pharmaverseadam)
library(admiral)
library(xportr)
library(dplyr)
library(lubridate)
library(stringr)
library(reactable)

# Read in input data
adsl <- pharmaverseadam::adsl
ae <- pharmaversesdtm::ae
ex <- pharmaversesdtm::ex

# When SAS datasets are imported into R using haven::read_sas(), missing
# character values from SAS appear as "" characters in R, instead of appearing
# as NA values. Further details can be obtained via the following link:
# https://pharmaverse.github.io/admiral/articles/admiral.html#handling-of-missing-values
ae <- convert_blanks_to_na(ae)
ex <- convert_blanks_to_na(ex)
```

# Load Specifications for Metacore {#loadspecs}

We have saved our specifications in an Excel file and will load them into `{metacore}` with the `metacore::spec_to_metacore()` function.

```{r echo=TRUE}
#| label: Load Specs
#| warning: false
# ---- Load Specs for Metacore ----
metacore <- spec_to_metacore(
  path = "./metadata/safety_specs.xlsx",
  # All datasets are described in the same sheet
  where_sep_sheet = FALSE
) %>%
  select_dataset("ADAE")
```


# Select ADSL Variables {#adslvars}

Some variables from the `ADSL` dataset required for the derivations are merged into the `AE` domain using the `admiral::derive_vars_merged()` function.
The rest of the relevant `ADSL` variables would be added later.

```{r}
# Select required ADSL variables
adsl_vars <- exprs(TRTSDT, TRTEDT, DTHDT)

# Join ADSL variables with VS
adae <- ae %>%
  derive_vars_merged(
    dataset_add = adsl,
    new_vars = adsl_vars,
    by_vars = exprs(STUDYID, USUBJID)
  )
```

# Start Building Derivations {#startbuild}

## Derive Analysis Dates {#anldates}

The first derivation step we are going to do is to compute the Analysis Date and Relative Analysis Day with the variables merged from `ADSL` dataset.

When deriving AE end date first as `AENDT`, we're going to impute partial dates with missing day or month.
For AE start date `ASTDT` then we're going to use the same imputation, but additionally we ensure that the imputed date doesn't go before treatment start date or after the AE end date.

```{r}
# Derive ASTDT/ASTDTF/ASTDY and AENDT/AENDTF/AENDY
adae <- adae %>%
  derive_vars_dt(
    new_vars_prefix = "AEN",
    dtc = AEENDTC,
    date_imputation = "last",
    highest_imputation = "M", # imputation is performed on missing days or months
    flag_imputation = "auto" # to automatically create AENDTF variable
  ) %>%
  derive_vars_dt(
    new_vars_prefix = "AST",
    dtc = AESTDTC,
    highest_imputation = "M", # imputation is performed on missing days or months
    flag_imputation = "auto", # to automatically create ASTDTF variable
    min_dates = exprs(TRTSDT), # apply a minimum date for the imputation
    max_dates = exprs(AENDT) # apply a maximum date for the imputation
  ) %>%
  derive_vars_dy(
    reference_date = TRTSDT,
    source_vars = exprs(ASTDT, AENDT)
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(adae %>% select(USUBJID, AESTDTC, ASTDT, ASTDTF, ASTDY, AEENDTC, AENDT, AENDTF, AENDY))
```

## Derive Duration {#anldur}

Now we have these date variables we can derive AE duration, where the default is to + 1 day, but this can be controlled with the `add_one` argument.

```{r}
# Derive ADURN/ADURU
adae <- adae %>%
  derive_vars_duration(
    new_var = ADURN,
    new_var_unit = ADURU,
    start_date = ASTDT,
    end_date = AENDT
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(adae %>% select(USUBJID, ASTDT, AENDT, ADURN, ADURU))
```

## Derive Date of Last Dose {#exdates}

You might need to add some exposure information from `ex` such as deriving the date of last dose before each AE.

In the below case we call this `LDOSEDT` and we want to look only at exposure records with a valid dose or placebo.

```{r}
# Derive LDOSEDT
# In our ex data the EXDOSFRQ (frequency) is "QD" which stands for once daily
# If this was not the case then we would need to use the admiral::create_single_dose_dataset() function
# to generate single doses from aggregate dose information
# Refer to https://pharmaverse.github.io/admiral/reference/create_single_dose_dataset.html
ex <- ex %>%
  derive_vars_dt(
    dtc = EXENDTC,
    new_vars_prefix = "EXEN"
  )

adae <- adae %>%
  derive_vars_joined(
    dataset_add = ex,
    by_vars = exprs(STUDYID, USUBJID),
    order = exprs(EXENDT),
    new_vars = exprs(LDOSEDT = EXENDT),
    join_vars = exprs(EXENDT),
    join_type = "all",
    filter_add = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO"))) & !is.na(EXENDT),
    filter_join = EXENDT <= ASTDT,
    mode = "last"
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(ex %>% select(USUBJID, EXTRT, EXDOSE, EXENDT))
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(adae %>% select(USUBJID, ASTDT, LDOSEDT))
```

## Derive Analysis Flags {anlflags}

Next we looks at common analysis flags such as treatment emergent and on treatment flags.
For the on-treatment flag we only want to include AEs occurring until 30 days after treatment end.

```{r}
# Derive TRTEMFL and ONTRTFL
adae <- adae %>%
  derive_var_trtemfl(
    start_date = ASTDT,
    end_date = AENDT,
    trt_start_date = TRTSDT,
    trt_end_date = TRTEDT
  ) %>%
  derive_var_ontrtfl(
    start_date = ASTDT,
    ref_start_date = TRTSDT,
    ref_end_date = TRTEDT,
    ref_end_window = 30
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(adae %>% select(USUBJID, ASTDT, TRTSDT, TRTEDT, TRTEMFL, ONTRTFL))
```

Both of the above `{admiral}` functions offer more flexibility if you needed to use more complex
analysis rules. For example, for treatment emergent you could use the intensity arguments if you
also wanted to flag those AEs starting before treatment start and ending after treatment start
with worsened intensity (i.e. the most extreme intensity is greater than the initial intensity).

## Derive Occurrence Flags {occflags}

There can be flags that need to be derived based on AE occurrences, such as flagging the first occurrence of maximum severity per patient which in the below example we call `AOCCIFL`.

```{r}
# Derive AOCCIFL
adae <- adae %>%
  # create temporary numeric ASEVN for sorting purpose
  mutate(TEMP_AESEVN = as.integer(factor(AESEV, levels = c("SEVERE", "MODERATE", "MILD")))) %>%
  derive_var_extreme_flag(
    new_var = AOCCIFL,
    by_vars = exprs(STUDYID, USUBJID),
    order = exprs(TEMP_AESEVN, ASTDT, AESEQ),
    mode = "first"
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(adae %>% select(USUBJID, ASTDT, AESEQ, AESEV, AOCCIFL))
```

## Derive Query Variables {query}

The final set of derived flags would be around those checking for queries/baskets of AE terms, e.g. using Standardized MedDRA Queries (SMQs) or Custom Queries (CQs). 

Before reading this section, you should read the [Queries Dataset Vignette](https://pharmaverse.github.io/admiral/articles/queries_dataset.html) from `{admiral}`.

On a study you would need to create your own version of this queries dataset with the AE queries you need, but for the purpose of this article we use the example provided as part of `{admiral}` and we filter it to just one CQ and one SMQ.

```{r}
queries <- admiral::queries %>%
  filter(PREFIX %in% c("CQ01", "SMQ02"))
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(queries)
```

```{r}
# Derive CQ01NAM and SMQ02NAM
adae <- adae %>%
  derive_vars_query(dataset_queries = queries)
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(adae %>% select(USUBJID, AEDECOD, CQ01NAM, SMQ02NAM))
```

## Add ADSL Variables {#adsladd}

If needed, the other `ADSL` variables can now be added.
List of ADSL variables already merged held in vector `adsl_vars`.

```{r eval=TRUE}
adae <- adae %>%
  derive_vars_merged(
    dataset_add = select(adsl, !!!negate_vars(adsl_vars)),
    by_vars = exprs(STUDYID, USUBJID)
  )
```

# Apply Metadata to Create an eSub XPT and Perform Associated Checks {#metacore_xportr}

We use `{metatools}` and `{xportr}` to perform checks, apply metadata such as types, lengths, labels, and write the dataset to an XPT file.

```{r checks, warning=FALSE, message=FALSE}
dir <- tempdir() # Specify the directory for saving the XPT file

adae %>%
  drop_unspec_vars(metacore) %>% # Drop unspecified variables from specs
  check_variables(metacore) %>% # Check all variables specified are present and no more
  check_ct_data(metacore, na_acceptable = TRUE) %>% # Checks all variables with CT only contain values within the CT
  order_cols(metacore) %>% # Orders the columns according to the spec
  sort_by_key(metacore) %>% # Sorts the rows by the sort keys
  xportr_type(metacore, domain = "ADAE") %>% # Coerce variable type to match spec
  xportr_length(metacore) %>% # Assigns SAS length from a variable level metadata
  xportr_label(metacore) %>% # Assigns variable label from metacore specifications
  xportr_df_label(metacore) %>%  # Assigns dataset label from metacore specifications
  xportr_write(file.path(dir, "adae.xpt"), metadata = metacore, domain = "ADAE")
```
