---
title: "ADER+"
order: 12
---

```{r setup script, include=FALSE, purl=FALSE}
invisible_hook_purl <- function(before, options, ...) {
  knitr::hook_purl(before, options, ...)
  NULL
}
knitr::knit_hooks$set(purl = invisible_hook_purl)
source("functions/print_df.R")
```

## Exposure-Response Analysis Data

# Introduction

Exposure-Response (ER) modeling is a critical tool in drug development for evaluating the relationships between drug exposure, safety, and efficacy. These analyses help characterize dose-response relationships, optimize dosing regimens, and support regulatory decision-making.

While CDISC released the Population Pharmacokinetic (PopPK) Implementation Guide in 2023 (see [Basic Data Structure for ADaM Population PK](https://www.cdisc.org/standards/foundational/adam/basic-data-structure-adam-poppk-implementation-guide-v1-0)), no equivalent standards currently exist specifically for ER data. However, ER datasets share many structural similarities with PopPK datasets, including numeric covariates, relative time variables, and pharmacokinetic exposure metrics. The examples below demonstrate how PopPK principles can be extended to create standardized ER analysis datasets.

The following four specialized datasets support different aspects of exposure-response analysis:

| Dataset | Full Name | Primary Purpose | Key Features |
|------------------|------------------|-------------------|------------------|
| [**ADER**](#ader) | Exposure-Response Analysis Dataset | Integrated exposure-response relationships across safety and efficacy endpoints | Combines PK metrics (AUC, Cmax) with time-to-event outcomes (OS, PFS), tumor response, and baseline covariates |
| [**ADEE**](#adee) | Exposure-Efficacy Analysis Dataset | Exposure relationships with efficacy and tumor response endpoints | Focuses on efficacy outcomes such as tumor measurements, response rates, and progression metrics linked to drug exposure |
| [**ADES**](#ades) | Exposure-Safety Analysis Dataset | Exposure relationships with adverse events and safety endpoints | Links exposure metrics to adverse event occurrence, severity, and time-to-onset for safety signal detection |
| [**ADTRR**](#adtrr) | Exposure-Tumor Response Rate Analysis Dataset | Exposure relationships with categorical tumor response metrics | Analyzes exposure impact on response categories (CR, PR, SD, PD) and best overall response |

Each dataset builds upon standard ADaM datasets (ADSL, ADRS, ADTTE, ADAE, ADLB, ADVS) and incorporates pharmacokinetic parameters from ADPC/ADPP to create analysis-ready datasets for exposure-response modeling.

The top of this page includes common derivations. See the tabs below for specifics for each dataset.

## First Load Packages

First we will load the packages required for our project. We will use `{admiral}` and `{admiralonco}` for the creation of analysis data. We will source these from `{pharmaverseadam}`. `{admiral}` requires `{dplyr}`, `{lubridate}` and `{stringr}`. Find other `{admiral}` functions and related variables by searching [admiraldiscovery](https://pharmaverse.github.io/admiraldiscovery/articles/reactable.html). We will use `{metacore}` and `{metatools}` to store and manipulate metadata from our specifications. We will use `{xportr}` to perform checks on the final data and export to a transport file.

Exposure Response data typically use ADaM data as source, so this example will depend on `{pharmaverseadam}` with data from existing `{admiral}` and `{admiralonco}` templates.

```{r echo=TRUE, message=FALSE}
#| label: Load Packages
# Load Packages
library(admiral)
library(admiralonco)
# pharmaverseadam contains example datasets generated from the CDISC pilot
# project SDTM ran through admiral templates
library(pharmaverseadam)
library(dplyr)
library(lubridate)
library(stringr)
library(metacore)
library(metatools)
library(xportr)
```

## Load Specifications for Metacore

We have saved our specifications in an Excel file and will load them into `{metacore}` with the `metacore::spec_to_metacore()` function. We will subset for `ADER` first and load the other datasets below.

```{r echo=TRUE, message=FALSE}
#| label: Load Specs
#| warning: false
# ---- Load Specs for Metacore ----
metacore <- spec_to_metacore("./metadata/pk_spec.xlsx") %>%
  select_dataset("ADER")
```

## Load Source Datasets

We will load ADaM data from `{pharmaverseadam}`. The main sources will be `ADRS`, `ADTTE`, `ADAE`, and `ADTR`. We will use `ADSL` for baseline characteristics and we will derive additional baselines from vital signs `ADVS` and laboratory data `ADLB`. We will use exposure metrics from `ADPP`.

```{r}
#| label: Load Source
# ---- Load source datasets ----
# Load ADRS, ADTTE, ADSL, ADLB, ADVS, ADEX, ADPP, ADAE and ADTR
adrs <- pharmaverseadam::adrs_onco
adtte <- pharmaverseadam::adtte_onco
adsl <- pharmaverseadam::adsl %>%
  filter(TRT01A != "Screen Failure")
adlb <- pharmaverseadam::adlb
advs <- pharmaverseadam::advs
adex <- pharmaverseadam::adex %>%
  filter(PARCAT1 == "INDIVIDUAL")
adpp <- pharmaverseadam::adpp
adae <- pharmaverseadam::adae
adtr <- pharmaverseadam::adtr_onco
```

## Common Derivations

We will include common derivations for all exposure-response datasets first. These include covariates and exposure metrics.

## Derive Covariates Using `{metatools}`

In this step we will create our numeric covariates using the `metatools::create_var_from_codelist()` function.

```{r}
#| label: Common Covariates
# ---- Derive Covariates ----
# Include numeric values for STUDYIDN, USUBJIDN, SEXN, RACEN etc.

covar <- adsl %>%
  create_var_from_codelist(metacore, input_var = STUDYID, out_var = STUDYIDN) %>%
  create_var_from_codelist(metacore, input_var = SEX, out_var = SEXN) %>%
  create_var_from_codelist(metacore, input_var = RACE, out_var = RACEN) %>%
  create_var_from_codelist(metacore, input_var = ETHNIC, out_var = ETHNICN) %>%
  create_var_from_codelist(metacore, input_var = ARMCD, out_var = COHORT) %>%
  create_var_from_codelist(metacore, input_var = ARMCD, out_var = COHORTC) %>%
  create_var_from_codelist(metacore, input_var = ARM, out_var = ARMN) %>%
  create_var_from_codelist(metacore, input_var = ACTARM, out_var = ACTARMN) %>%
  create_var_from_codelist(metacore, input_var = TRT01A, out_var = TRT01AN) %>%
  create_var_from_codelist(metacore, input_var = TRT01P, out_var = TRT01PN) %>%
  create_var_from_codelist(metacore, input_var = COUNTRY, out_var = COUNTRYN) %>%
  create_var_from_codelist(metacore, input_var = COUNTRY, out_var = COUNTRYL) %>%
  mutate(
    STUDYIDN = as.numeric(word(USUBJID, 1, sep = fixed("-"))),
    SITEIDN = as.numeric(word(USUBJID, 2, sep = fixed("-"))),
    USUBJIDN = as.numeric(word(USUBJID, 3, sep = fixed("-"))),
    SUBJIDN = as.numeric(SUBJID),
    ROUTE = unique(adex$EXROUTE)[1],
    FORM = unique(adex$EXDOSFRM)[1],
    REGION1 = COUNTRY,
    REGION1N = COUNTRYN,
  ) %>%
  create_var_from_codelist(metacore, input_var = FORM, out_var = FORMN) %>%
  create_var_from_codelist(metacore, input_var = ROUTE, out_var = ROUTEN)
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print covar
print_df(covar %>% select(USUBJID, TRT01A, TRT01AN, SEX, SEXN, RACE, RACEN, FORM, FORMN))
```

### Derive Additional Baselines

Next we add additional baselines from vital signs and laboratory data.

```{r}
#| label: Baselines
# ---- Derive additional baselines from ADVS and ADLB ----

labsbl <- adlb %>%
  filter(ABLFL == "Y" & PARAMCD %in% c("CREAT", "ALT", "AST", "BILI")) %>%
  mutate(PARAMCDB = paste0(PARAMCD, "BL")) %>%
  select(STUDYID, USUBJID, PARAMCDB, AVAL)

covar_vslb <- covar %>%
  derive_vars_merged(
    dataset_add = advs,
    filter_add = PARAMCD == "HEIGHT" & ABLFL == "Y",
    by_vars = exprs(STUDYID, USUBJID),
    new_vars = exprs(HTBL = AVAL)
  ) %>%
  derive_vars_merged(
    dataset_add = advs,
    filter_add = PARAMCD == "WEIGHT" & ABLFL == "Y",
    by_vars = exprs(STUDYID, USUBJID),
    new_vars = exprs(WTBL = AVAL)
  ) %>%
  derive_vars_transposed(
    dataset_merge = labsbl,
    by_vars = exprs(STUDYID, USUBJID),
    key_var = PARAMCDB,
    value_var = AVAL
  ) %>%
  mutate(
    BMIBL = compute_bmi(height = HTBL, weight = WTBL),
    BSABL = compute_bsa(
      height = HTBL,
      weight = WTBL,
      method = "Mosteller"
    ),
    CRCLBL = compute_egfr(
      creat = CREATBL, creatu = "SI", age = AGE, weight = WTBL, sex = SEX,
      method = "CRCL"
    ),
    EGFRBL = compute_egfr(
      creat = CREATBL, creatu = "SI", age = AGE, weight = WTBL, sex = SEX,
      method = "CKD-EPI"
    )
  ) %>%
  rename(TBILBL = BILIBL)
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print covar_vslb
print_df(covar_vslb %>% select(USUBJID, WTBL, HTBL, BMIBL, CRCLBL))
```

## Add Exposure Metrics from `ADPP`

An essential component of ER modeling is the exposure metrics calculated from the PK concentration data in `ADPC` or `ADPP`. Here we will use `AUCLST` and `CMAX` from the `ADPP` dataset from `{admiral}` in `{pharmaverseadam}`. See the `{aNCA}` package for details about calculating `AUC`, `CMAX`, and other parameters.

```{r}
#| label: AUCSS
# ---- Add Exposure Metrics  ----
covar_auc <- covar_vslb %>%
  derive_vars_transposed(
    dataset_merge = adpp,
    filter = PARAMCD %in% c("AUCLST", "CMAX"),
    by_vars = get_admiral_option("subject_keys"),
    key_var = PARAMCD,
    value_var = AVAL
  ) %>%
  rename(AUCSS = AUCLST, CMAXSS = CMAX)
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print covar_auc
print_df(covar_auc %>% select(
  USUBJID, TRT01A, SEX, WTBL, HTBL, AUCSS, CMAXSS
))
```

# Code for Individual Datasets

Derivations specific to each dataset continue below. Select the appropriate tab to view.

## Dataset-Specific Derivations

::: {.panel-tabset}

## ADER {#ader}

### Derive Time to Event Variables from `ADTTE`

We will use parameters from from `ADTTE` for Overall Survival, Progression Free Survival and Duration of Response. We will use `admiral::derive_vars_transposed()` to transpose the `AVAL` for each `PARAMCD`.

```{r}
#| label: ADER Add ADTTE variables
# ---- Derivations ----
# For ADTTE censor variables add "IND" to PARAMCD
adttei <- adtte %>%
  mutate(PARAMCD = paste0(PARAMCD, "IND"))

ader_tte <- adsl %>%
  select(!!!get_admiral_option("subject_keys")) %>%
  # Create OS and PFS variables from ADTTE
  derive_vars_transposed(
    dataset_merge = adtte,
    by_vars = get_admiral_option("subject_keys"),
    key_var = PARAMCD,
    value_var = AVAL
  ) %>%
  # Create OS and PFS censor variables
  derive_vars_transposed(
    dataset_merge = adttei,
    by_vars = get_admiral_option("subject_keys"),
    key_var = PARAMCD,
    value_var = CNSR
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print ader_tte
print_df(ader_tte %>% select(USUBJID, OS, OSIND, PFS, PFSIND, RSD, RSDIND))
```

We will use Best Overall Response by Investigator (BOR) from `ADRS`. We attach this to the data with `admiral::derive_vars_merged()`. Note that `derive_vars_transposed()` and `derive_vars_merged()` will be our primary ways of combining data from the different `{admiral}` and `{admiralonco}` templates.

```{r}
#| label: ADER Add ADRS
# ---- Add ADRS data ----
# Add response date to ADSL for duration of response calculation
ader_bor <- ader_tte %>%
  derive_vars_merged(
    dataset_add = adrs,
    filter_add = PARAMCD == "BOR" & ANL01FL == "Y",
    by_vars = get_admiral_option("subject_keys"),
    new_vars = exprs(BOR = AVAL, BORC = AVALC)
  )
```

Add Analysis Sequence Number using `admiral::derive_var_obs_number()`.

```{r}
#| label: ADER ASEQ
# ---- Add Sequence Number ----
ader_aseq <- ader_bor %>%
  derive_var_obs_number(
    by_vars = get_admiral_option("subject_keys"),
    check_type = "error"
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print ader_bor
print_df(ader_bor %>% select(
  USUBJID, BOR, BORC
))
```

### Combine with Covariates

We combine our covariates with the rest of the data

```{r}
#| label: ADER Combine with Covariates
# Combine covariates with ADER data

ader_prefinal <- ader_aseq %>%
  derive_vars_merged(
    dataset_add = covar_auc,
    by_vars = exprs(STUDYID, USUBJID)
  )
```

### Check Data With metacore and metatools

We use `{metacore}` objects with `{metatools}` functions to perform a number of checks on the data. We will drop variables not in the specs and make sure all the variables from the specs are included.

```{r}
#| label: ADER Metacore
#| warning: false

ader <- ader_prefinal %>%
  drop_unspec_vars(metacore) %>% # Drop unspecified variables from specs
  check_variables(metacore) %>% # Check all variables specified are present and no more
  check_ct_data(metacore) %>% # Checks all variables with CT only contain values within the CT
  order_cols(metacore) %>% # Orders the columns according to the spec
  sort_by_key(metacore) # Sorts the rows by the sort keys
```

### Apply Labels and Formats with xportr

Using `{xportr}` we check variable type, assign variable length, add variable labels, add variable formats, and save a transport file with `xportr::xportr_write()`.

```{r}
#| label: ADER xportr
dir <- tempdir() # Change to whichever directory you want to save the dataset in

ader_xpt <- ader %>%
  xportr_type(metacore, domain = "ADER") %>% # Coerce variable type to match spec
  xportr_length(metacore) %>% # Assigns SAS length from a variable level metadata
  xportr_label(metacore) %>% # Assigns variable label from metacore specifications
  xportr_format(metacore) %>% # Assigns variable format from metacore specifications
  xportr_df_label(metacore) %>% # Assigns dataset label from metacore specifications
  xportr_write(file.path(dir, "ader.xpt")) # Write xpt v5 transport file
```

## ADEE {#adee}

### Load Specifications

```{r echo=TRUE, message=FALSE}
#| label: ADEE Load Specs
#| warning: false
metacore <- spec_to_metacore("./metadata/pk_spec.xlsx") %>%
  select_dataset("ADEE")
```

### Create Base `ADEE` dataset from `ADTTE`

```{r}
#| label: ADEE Base dataset
# ---- Create adee base dataset

# Get variable names from both datasets
adsl_vars <- names(adsl)
adtte_vars <- names(adtte)

# Find common variables
common_vars <- intersect(adsl_vars, adtte_vars)

# Remove key variables to get variables to drop
vars_to_drop <- setdiff(common_vars, c("STUDYID", "USUBJID"))

# Ensure PARAMN exists in ADTTE
if (!"PARAMN" %in% names(adtte)) {
  adtte <- adtte %>%
    mutate(
      PARAMN = case_when(
        PARAMCD == "PFS" ~ 1,
        PARAMCD == "OS" ~ 2,
        PARAMCD == "TTP" ~ 3,
        PARAMCD == "TTNT" ~ 4,
        TRUE ~ 99
      )
    )
}

# ---- Create ADEE Base

adee_base <- adtte %>%
  # Filter to efficacy endpoints
  filter(PARAMCD %in% c("OS", "PFS", "TTP", "TTNT")) %>%
  # Add derived variables
  mutate(
    EVENT = 1 - CNSR,
    AVALU = if_else(!is.na(AVAL), "DAYS", NA_character_),
  ) %>%
  # Remove overlapping variables (use clean method)
  select(-any_of(vars_to_drop))
```


```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print adee_base
print_df(adee_base %>% select(USUBJID, PARAMCD, PARAM, AVAL, AVALU))
```


### Add Analysis Variables

```{r}
#| label: ADEE Analysis variables
# ---- Add Analysis variables

adee_aseq <- adee_base %>%
  # Analysis flags
  mutate(
    ANL01FL = if_else(PARAMCD == "PFS", "Y", ""),
    ANL02FL = if_else(PARAMCD == "OS", "Y", ""),
    ANL03FL = if_else(PARAMCD == "TTP", "Y", ""),
    ANL04FL = if_else(PARAMCD == "TTNT", "Y", "")
  ) %>%
  # Parameter categories
  mutate(
    PARCAT1 = "EFFICACY",
    PARCAT2 = "TIME TO EVENT"
  ) %>%
  # Sequence number
  derive_var_obs_number(
    by_vars = exprs(STUDYID, USUBJID),
    order = exprs(PARAMCD),
    new_var = ASEQ,
    check_type = "error"
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print adee_seq
print_df(adee_aseq %>% select(USUBJID, PARAMCD, PARAM, AVAL, AVALU, ANL01FL, ANL02FL, ASEQ))
```


### Combine with Covariates and Exposure

We combine our covariates with the rest of the data

```{r}
#| label: ADEE Combine with Covariates
# Combine covariates with ADEE data

adee_prefinal <- adee_aseq %>%
  derive_vars_merged(
    dataset_add = covar_auc,
    by_vars = exprs(STUDYID, USUBJID)
  )
```



```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print adee_prefinal
print_df(adee_prefinal %>% select(USUBJID, TRT01PN, AUCSS, CMAXSS, PARAMCD, AVAL, AVALU), n = 20)
```

### Check Data With metacore and metatools

We use `{metacore}` objects with `{metatools}` functions to perform a number of checks on the data. We will drop variables not in the specs and make sure all the variables from the specs are included.

```{r}
#| label: ADEE Metacore
#| warning: false

adee <- adee_prefinal %>%
  drop_unspec_vars(metacore) %>% # Drop unspecified variables from specs
  check_variables(metacore) %>% # Check all variables specified are present and no more
  check_ct_data(metacore) %>% # Checks all variables with CT only contain values within the CT
  order_cols(metacore) %>% # Orders the columns according to the spec
  sort_by_key(metacore) # Sorts the rows by the sort keys
```

### Apply Labels and Formats with xportr

Using {xportr} we check variable type, assign variable length, add variable labels, add variable formats, and save a transport file with `xportr::xportr_write()`.

```{r}
#| label: ADEE xportr
dir <- tempdir() # Change to whichever directory you want to save the dataset in

adee_xpt <- adee %>%
  xportr_type(metacore, domain = "ADEE") %>% # Coerce variable type to match spec
  xportr_length(metacore) %>% # Assigns SAS length from a variable level metadata
  xportr_label(metacore) %>% # Assigns variable label from metacore specifications
  xportr_format(metacore) %>% # Assigns variable format from metacore specifications
  xportr_df_label(metacore) %>% # Assigns dataset label from metacore specifications
  xportr_write(file.path(dir, "adee.xpt")) # Write xpt v5 transport file
```

## ADES {#ades}

### Load Specifications

```{r echo=TRUE, message=FALSE}
#| label: ADES Load Specs
#| warning: false
metacore <- spec_to_metacore("./metadata/pk_spec.xlsx") %>%
  select_dataset("ADES")
```

### Create `ADES` base dataset

```{r}
#| label: ADES Base data
# ---- Create ades base dataset

# Derive subject-level summary parameters from ADAE
# NOTE: pharmaverseadam uses ASEV/ASEVN (severity) not AETOXGR/AETOXGRN

# Get all subjects from ADSL

adsl_sub <- adsl %>%
  select(!!!get_admiral_option("subject_keys"))


## Any TEAE (Treatment-Emergent Adverse Event) ----
teae_any <- adsl_sub %>%
  mutate(
    PARAMCD = "TEAE",
    PARAM = "Any Treatment-Emergent Adverse Event",
    PARAMN = 1,
    AVAL = if_else(
      USUBJID %in% (adae %>% filter(TRTEMFL == "Y") %>% pull(USUBJID)),
      1, 0
    ),
    AVALC = if_else(AVAL == 1, "Y", "N")
  )

## Any Severe TEAE (using ASEVN = 3 for SEVERE) ----
teae_sev <- adsl_sub %>%
  mutate(
    PARAMCD = "TEAESEV", # Using ASEVN instead of AETOXGRN
    PARAM = "Any Severe Treatment-Emergent Adverse Event",
    PARAMN = 2,
    AVAL = if_else(
      USUBJID %in% (adae %>% filter(TRTEMFL == "Y" & ASEVN == 3) %>% pull(USUBJID)),
      1, 0
    ),
    AVALC = if_else(AVAL == 1, "Y", "N")
  )

## Any Serious AE ----
sae_any <- adsl_sub %>%
  mutate(
    PARAMCD = "SAE",
    PARAM = "Any Serious Adverse Event",
    PARAMN = 3,
    AVAL = if_else(
      USUBJID %in% (adae %>% filter(AESER == "Y") %>% pull(USUBJID)),
      1, 0
    ),
    AVALC = if_else(AVAL == 1, "Y", "N")
  )

## Any Treatment-Related AE (using AREL character values) ----
trae_any <- adsl_sub %>%
  mutate(
    PARAMCD = "TRAE",
    PARAM = "Any Treatment-Related Adverse Event",
    PARAMN = 4,
    AVAL = if_else(
      USUBJID %in% (adae %>%
        filter(AREL %in% c("POSSIBLE", "PROBABLE", "RELATED")) %>%
        pull(USUBJID)),
      1, 0
    ),
    AVALC = if_else(AVAL == 1, "Y", "N")
  )

## AE Leading to Treatment Discontinuation ----
aedcn <- adsl_sub %>%
  mutate(
    PARAMCD = "AEDCN",
    PARAM = "AE Leading to Treatment Discontinuation",
    PARAMN = 5,
    AVAL = if_else(
      USUBJID %in% (adae %>% filter(AEACN == "DRUG WITHDRAWN") %>% pull(USUBJID)),
      1, 0
    ),
    AVALC = if_else(AVAL == 1, "Y", "N")
  )

# Combine subject-level parameters
subject_params <- bind_rows(
  teae_any,
  teae_sev,
  sae_any,
  trae_any,
  aedcn
) %>%
  arrange(USUBJID, PARAMCD)
```


```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print subject_params
print_df(subject_params %>% select(USUBJID, PARAMCD, PARAM, AVAL, AVALC), n = 20)
```


### Create Event Level Parameters

```{r}
#| label: ADES Event level parameters

# ---- Create event level parameters

# Get variable names for clean dropping
adsl_vars <- names(covar_auc)
adae_vars <- names(adae)
common_vars <- intersect(adsl_vars, adae_vars)
vars_to_drop <- setdiff(common_vars, c("STUDYID", "USUBJID"))

# Create event-level records from ADAE
# NOTE: Using actual pharmaverseadam variables (ASEV/ASEVN, AREL)
event_params <- adae %>%
  filter(TRTEMFL == "Y") %>% # Treatment-emergent only
  mutate(
    PARAMCD = "AETERM",
    PARAM = "Adverse Event Term",
    PARAMN = 10,
    AVAL = 1, # Event occurred
    AVALC = "Y",

    # Keep AE-specific variables (8-char names)
    # Using actual pharmaverseadam ADAE variables
    AEDECOD = AEDECOD, # Preferred term
    AEBODSYS = AEBODSYS, # System organ class
    AESEV = ASEV, # Severity (char): MILD, MODERATE, SEVERE
    AESEVN = ASEVN, # Severity (num): 1, 2, 3
    AESER = AESER, # Serious flag: Y/N
    AEREL = AREL, # Relationship (char): NOT RELATED, POSSIBLE, etc.

    # Create numeric relationship for analysis
    AERELN = case_when(
      AREL == "NOT RELATED" ~ 0,
      AREL == "UNLIKELY RELATED" ~ 1,
      AREL == "POSSIBLE" ~ 2,
      AREL == "PROBABLE" ~ 3,
      AREL == "RELATED" ~ 4,
      TRUE ~ NA_real_
    ),
    AESTDT = ASTDT, # AE start date (8 chars)
    AEENDT = AENDT # AE end date (8 chars)
  ) %>%
  select(-any_of(vars_to_drop))

# ---- Combine subject and event levels

# Ensure all AE-specific variables exist in subject_params (as NA)
# This prevents issues when binding with event_params
subject_params_complete <- subject_params %>%
  derive_vars_merged(
    dataset_add = adsl_sub,
    by_vars = exprs(STUDYID, USUBJID),
  ) %>%
  mutate(
    # Add event-level variables as NA for subject-level records
    AESTDT = as.Date(NA),
    AEENDT = as.Date(NA),
    AEDECOD = NA_character_,
    AEBODSYS = NA_character_,
    AESEV = NA_character_,
    AESEVN = NA_integer_,
    AESER = NA_character_,
    AEREL = NA_character_,
    AERELN = NA_real_
  )

event_params_complete <- event_params %>%
  derive_vars_merged(
    dataset_add = adsl_sub,
    by_vars = exprs(STUDYID, USUBJID)
  )

# Combine both levels
ades_base <- bind_rows(
  subject_params_complete,
  event_params_complete
) %>%
  arrange(USUBJID, PARAMCD)
```


```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print ades_base
print_df(ades_base %>% select(USUBJID, PARAMCD, PARAM, AEDECOD, AVAL, AVALC, AESTDY), n = 20)
```


### Add Analysis Variables and Flags

```{r}
#| label: ADES analysis variables
# ---- Add analysis variables

ades_flags <- ades_base %>%
  # Analysis flags
  mutate(
    ANL01FL = if_else(PARAMCD == "TEAE", "Y", ""),
    ANL02FL = if_else(PARAMCD == "TEAESEV", "Y", ""),
    ANL03FL = if_else(PARAMCD == "SAE", "Y", ""),
    ANL04FL = if_else(PARAMCD == "TRAE", "Y", ""),
    ANL05FL = if_else(PARAMCD == "AETERM", "Y", "")
  ) %>%
  # Parameter categories
  mutate(
    PARCAT1 = "SAFETY",
    PARCAT2 = case_when(
      PARAMN <= 5 ~ "SUBJECT-LEVEL",
      PARAMN >= 10 ~ "EVENT-LEVEL",
      TRUE ~ NA_character_
    )
  ) %>%
  # Analysis timepoint
  mutate(
    AVISIT = if_else(PARAMN <= 5, "OVERALL", "AT EVENT"),
    AVISITN = if_else(PARAMN <= 5, 99, 0)
  ) %>%
  # Sort and create sequence number
  # Use coalesce to handle NA dates (puts them first in sort)
  arrange(USUBJID, PARAMN, coalesce(AESTDT, as.Date("1900-01-01"))) %>%
  group_by(STUDYID, USUBJID) %>%
  mutate(ASEQ = row_number()) %>%
  ungroup()
```

### Combine with Covariates and Exposure

We combine our covariates with the rest of the data

```{r}
#| label: ADES Combine with Covariates
# Combine covariates with ADES data

ades_prefinal <- ades_flags %>%
  derive_vars_merged(
    dataset_add = covar_auc,
    by_vars = exprs(STUDYID, USUBJID)
  )
```


### Check Data With metacore and metatools

```{r}
#| label: ADES metatools
## Check Data With metacore and metatools

ades <- ades_prefinal %>%
  drop_unspec_vars(metacore) %>% # Drop unspecified variables from specs
  check_variables(metacore) %>% # Check all variables specified are present and no more
  check_ct_data(metacore) %>% # Checks all variables with CT only contain values within the CT
  order_cols(metacore) %>% # Orders the columns according to the spec
  sort_by_key(metacore) # Sorts the rows by the sort keys
```

### Apply Labels and Formats with xportr

Using {xportr} we check variable type, assign variable length, add variable labels, add variable formats, and save a transport file with `xportr::xportr_write()`.

```{r}
#| label: ADES xportr

dir <- tempdir() # Change to whichever directory you want to save the dataset in

ades_xpt <- ades %>%
  xportr_type(metacore, domain = "ADES") %>% # Coerce variable type to match spec
  xportr_length(metacore) %>% # Assigns SAS length from a variable level metadata
  xportr_label(metacore) %>% # Assigns variable label from metacore specifications
  xportr_format(metacore) %>% # Assigns variable format from metacore specifications
  xportr_df_label(metacore, domain = "ADES") %>% # Assigns dataset label from metacore specifications
  xportr_write(file.path(dir, "ades.xpt")) # Write xpt v5 transport file
```


## ADTRR {#adtrr}

### Load Specifications for `ADTRR`

```{r echo=TRUE, message=FALSE}
#| label: ADTRR Load Specs
#| warning: false
metacore <- spec_to_metacore("./metadata/pk_spec.xlsx") %>%
  select_dataset("ADTRR")
```

### Create `ADTRR` Base Data

```{r}
#| label: ADTRR Base Data

# ---- Create base tumor size parameter dataset

# Get variable names for clean dropping
adsl_vars <- names(covar_auc)
adtr_vars <- names(adtr)
common_vars <- intersect(adsl_vars, adtr_vars)
vars_to_drop <- setdiff(common_vars, c("STUDYID", "USUBJID"))

tsize_final <- adtr %>%
  filter(PARAMCD == "SDIAM") %>%
  mutate(
    PARAMCD = "TSIZE",
    PARAM = "Target Lesions Sum of Diameters",
    PARAMN = 1
  ) %>%
  derive_var_nfrlt(
    new_var = NFRLT,
    new_var_unit = FRLTU,
    out_unit = "HOURS",
    visit_day = ADY
  ) %>%
  select(-any_of(vars_to_drop))
```


```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(tsize_final %>% select(USUBJID, PARAMCD, PARAM, AVISIT, NFRLT, BASE, AVAL, CHG), n = 20)
```



### Add BOR from `ADRS`

```{r}
#| label: ADTRR add BOR

# ---- Add BOR from ADRS

adrs_vars <- names(adrs)
common_vars_adrs <- intersect(adsl_vars, adrs_vars)
vars_to_drop_adrs <- setdiff(common_vars_adrs, c("STUDYID", "USUBJID"))

bor <- adrs %>%
  filter(PARAMCD == "BOR" & SAFFL == "Y") %>%
  mutate(
    PARAMN = 2,
    # Create BORN from AVALC if AVAL doesn't exist
    BORN = if ("AVAL" %in% names(.)) {
      AVAL
    } else {
      case_when(
        AVALC == "CR" ~ 4,
        AVALC == "PR" ~ 3,
        AVALC == "SD" ~ 2,
        AVALC == "PD" ~ 1,
        AVALC == "NE" ~ 0,
        TRUE ~ NA_real_
      )
    }
  ) %>%
  select(-any_of(vars_to_drop_adrs))
```



```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print bor
print_df(bor %>% select(USUBJID, PARAMCD, PARAM, AVAL, AVALC), n = 20)
```


### Derive Nadir

```{r}
#| label: ADTRR add NADIR
# ---- Derive Nadir

# Calculate nadir from TSIZE records
# Keep BASE, CHG, PCHG from the nadir timepoint
nadir <- tsize_final %>%
  filter(AVISITN > 0 & !is.na(AVAL)) %>%
  group_by(STUDYID, USUBJID) %>%
  filter(AVAL == min(AVAL, na.rm = TRUE)) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(
    PARAMCD = "NADIR",
    PARAM = "Nadir Tumor Size",
    PARAMN = 3,
    NADIR = AVAL,
    NADPCHG = PCHG, # Keep PCHG at nadir
    NADVST = AVISIT # Keep visit of nadir
  )
```


```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print nadir
print_df(nadir %>% select(USUBJID, PARAMCD, PARAM, NADIR, NADPCHG), n = 20)
```



### Combine Parameters

```{r}
#| label: ADTRR combine parameters

# ---- Combine parameters

adtrr_base <- bind_rows(
  tsize_final,
  bor,
  nadir
) %>%
  arrange(USUBJID, PARAMN, AVISITN)

# ---- Add analysis variables

# Ensure AVALU exists before mutating
if (!"AVALU" %in% names(adtrr_base)) {
  adtrr_base <- adtrr_base %>%
    mutate(AVALU = NA_character_)
}

adtrr_seq <- adtrr_base %>%
  # Analysis flags
  mutate(
    # Baseline flag
    ABLFL = case_when(
      !is.na(ABLFL) ~ ABLFL,
      !is.na(AVISITN) & AVISITN == 0 ~ "Y",
      TRUE ~ ""
    ),

    # Post-baseline flag
    ANL01FL = if_else(!is.na(AVISITN) & AVISITN > 0, "Y", ""),

    # Responders (CR or PR)
    ANL02FL = if_else(!is.na(AVALC) & AVALC %in% c("CR", "PR"), "Y", ""),

    # Has change from baseline
    ANL03FL = if_else(!is.na(PCHG), "Y", "")
  ) %>%
  # Parameter categories
  mutate(
    PARCAT1 = "TUMOR RESPONSE",
    PARCAT2 = case_when(
      PARAMCD == "TSIZE" ~ "MEASUREMENT",
      PARAMCD == "BOR" ~ "OVERALL RESPONSE",
      PARAMCD == "NADIR" ~ "BEST RESPONSE",
      TRUE ~ NA_character_
    )
  ) %>%
  # Set AVALU (now guaranteed to exist)
  mutate(
    AVALU = case_when(
      !is.na(AVALU) & AVALU != "" ~ AVALU, # Keep existing non-empty
      PARAMCD == "TSIZE" ~ "mm",
      PARAMCD == "NADIR" ~ "mm",
      TRUE ~ NA_character_
    )
  ) %>%
  # Sequence number
  derive_var_obs_number(
    by_vars = exprs(STUDYID, USUBJID),
    order = exprs(PARAMN, AVISITN),
    new_var = ASEQ,
    check_type = "error"
  ) %>%
  arrange(USUBJID, PARAMN, AVISITN)
```



```{r eval=TRUE, echo=FALSE, purl=FALSE}
#| label: print adtrr_seq
print_df(adtrr_seq %>% select(USUBJID, PARAMCD, PARAM, AVISIT, NFRLT, BASE, AVAL, CHG), n = 20)
```


### Combine with Covariate and Exposure Data

```{r}
#| label: ADTRR prefinal
# ---- Combine with covariates

adtrr_prefinal <- adtrr_seq %>%
  derive_vars_merged(
    dataset_add = covar_auc,
    by_vars = exprs(STUDYID, USUBJID)
  )
```

### Check Data With metacore and metatools

```{r}
#| label: ADTTR metatools
## Check Data With metacore and metatools

adtrr <- adtrr_prefinal %>%
  drop_unspec_vars(metacore) %>% # Drop unspecified variables from specs
  check_variables(metacore) %>% # Check all variables specified are present and no more
  check_ct_data(metacore) %>% # Checks all variables with CT only contain values within the CT
  order_cols(metacore) %>% # Orders the columns according to the spec
  sort_by_key(metacore) # Sorts the rows by the sort keys
```

### Apply Labels and Formats with xportr

Using {xportr} we check variable type, assign variable length, add variable labels, add variable formats, and save a transport file with `xportr::xportr_write()`.

```{r}
#| label: ADTRR xportr

dir <- tempdir() # Change to whichever directory you want to save the dataset in

adtrr_xpt <- adtrr %>%
  xportr_type(metacore, domain = "ADTRR") %>% # Coerce variable type to match spec
  xportr_length(metacore) %>% # Assigns SAS length from a variable level metadata
  xportr_label(metacore) %>% # Assigns variable label from metacore specifications
  xportr_format(metacore) %>% # Assigns variable format from metacore specifications
  xportr_df_label(metacore, domain = "ADTRR") %>% # Assigns dataset label from metacore specifications
  xportr_write(file.path(dir, "adtrr.xpt")) # Write xpt v5 transport file
```


:::

# Conclusions

These Exposure-Response datasets demonstrate a practical approach to standardizing ER data using CDISC ADaM principles and the Pharmaverse ecosystem. The four datasets work together to support different modeling objectives: `ADER` for integrated exposure-response analysis, `ADEE` for efficacy endpoints, `ADES` for safety evaluation, and `ADTRR` for tumor response assessment. The framework demonstrated here can be adapted for study-specific ER endpoints. By standardizing the structure of exposure-response data and maintaining consistency with other ADaM formats, these datasets facilitate regulatory review, enhance traceability, and enable reproducible pharmacometric analyses. We encourage the community to adopt and refine these approaches as CDISC continues to develop formal ER data standards.



