---
title: "Exposure-Response"
order: 12
---

```{r setup script, include=FALSE, purl=FALSE}
invisible_hook_purl <- function(before, options, ...) {
  knitr::hook_purl(before, options, ...)
  NULL
}
knitr::knit_hooks$set(purl = invisible_hook_purl)
source("functions/print_df.R")
```

# Introduction

Exposure-Response (ER) modeling is a critical tool in drug development for evaluating the relationships between drug exposure, safety, and efficacy. These analyses help characterize dose-response relationships, optimize dosing regimens, and support regulatory decision-making.

While CDISC released the Population Pharmacokinetic (PopPK) Implementation Guide in 2023 (see [Basic Data Structure for ADaM Population PK](https://www.cdisc.org/standards/foundational/adam/basic-data-structure-adam-poppk-implementation-guide-v1-0)), no equivalent standards currently exist specifically for ER data. However, ER datasets share many structural similarities with PopPK datasets, including numeric covariates, relative time variables, and pharmacokinetic exposure metrics. The examples below demonstrate how PopPK principles can be extended to create standardized ER analysis datasets.

The following four specialized datasets support different aspects of exposure-response analysis:

| Dataset | Full Name | Primary Purpose | Key Features |
|------------------|------------------|-------------------|------------------|
| **[ADER](#ader)** | Exposure-Response Analysis Dataset | Integrated exposure-response relationships across safety and efficacy endpoints | Combines PK metrics (AUC, Cmax) with time-to-event outcomes (OS, PFS), tumor response, and baseline covariates |
| **[ADEE](#adee)** | Exposure-Efficacy Analysis Dataset | Exposure relationships with efficacy and tumor response endpoints | Focuses on efficacy outcomes such as tumor measurements, response rates, and progression metrics linked to drug exposure |
| **[ADES](#ades)** | Exposure-Safety Analysis Dataset | Exposure relationships with adverse events and safety endpoints | Links exposure metrics to adverse event occurrence, severity, and time-to-onset for safety signal detection |
| **[ADTRR](#adtrr)** | Exposure-Tumor Response Rate Analysis Dataset | Exposure relationships with categorical tumor response metrics | Analyzes exposure impact on response categories (CR, PR, SD, PD) and best overall response |

Each dataset builds upon standard ADaM datasets (ADSL, ADRS, ADTTE, ADAE, ADLB, ADVS) and incorporates pharmacokinetic parameters from ADPC/ADPP to create analysis-ready datasets for exposure-response modeling.

## First Load Packages

First we will load the packages required for our project. We will use `{admiral}` and `{admiralonco}` for the creation of analysis data. We will source these from `{pharmaverseadam}`. `{admiral}` requires `{dplyr}`, `{lubridate}` and `{stringr}`. Find other `{admiral}` functions and related variables by searching [admiraldiscovery](https://pharmaverse.github.io/admiraldiscovery/articles/reactable.html). We will use `{metacore}` and `{metatools}` to store and manipulate metadata from our specifications. We will use `{xportr}` to perform checks on the final data and export to a transport file.

Exposure Response data typically use ADaM data as source, so this example will depend on `{pharmaverseadam}` with data from existing `{admiral}` and `{admiralonco}` templates.

```{r echo=TRUE, message=FALSE}
#| label: Load Packages
# Load Packages
library(admiral)
library(admiralonco)
# pharmaverseadam contains example datasets generated from the CDISC pilot
# project SDTM ran through admiral templates
library(pharmaverseadam)
library(dplyr)
library(lubridate)
library(stringr)
library(metacore)
library(metatools)
library(xportr)
```

## Next Load Specifications for Metacore

We have saved our specifications in an Excel file and will load them into `{metacore}` with the `metacore::spec_to_metacore()` function.

```{r echo=TRUE, message=FALSE}
#| label: Load Specs
#| warning: false
# ---- Load Specs for Metacore ----
metacore <- spec_to_metacore("./metadata/pk_spec.xlsx") %>%
  select_dataset("ADER")
```

## Load Source Datasets

We will load are ADaM data from `{pharmaverseadam}`. The main source will be `ADRS` and `ADTTE`. We will use `ADSL` for baseline characteristics and we will derive additional baselines from vital signs `ADVS` and laboratory data `ADLB`.

```{r}
#| label: Load Source
# ---- Load source datasets ----
# Load ADRS, ADTTE, ADSL, ADLB, ADVS, ADEX, ADPP and ADAE
adrs <- pharmaverseadam::adrs_onco
adtte <- pharmaverseadam::adtte_onco
adsl <- pharmaverseadam::adsl
adlb <- pharmaverseadam::adlb
advs <- pharmaverseadam::advs
adex <- pharmaverseadam::adex %>%
  filter(PARCAT1 == "INDIVIDUAL")
adpp <- pharmaverseadam::adpp
adae <- pharmaverseadam::adae
```

## Common Derivations

We will include common derivations for all exposure-response datasets first. These include covariates and exposure metrics.

## Derive Covariates Using `{metatools}`

In this step we will create our numeric covariates using the `metatools::create_var_from_codelist()` function.

```{r}
#| label: Common Covariates
# ---- Derive Covariates ----
# Include numeric values for STUDYIDN, USUBJIDN, SEXN, RACEN etc.

covar <- adsl %>%
  create_var_from_codelist(metacore, input_var = STUDYID, out_var = STUDYIDN) %>%
  create_var_from_codelist(metacore, input_var = SEX, out_var = SEXN) %>%
  create_var_from_codelist(metacore, input_var = RACE, out_var = RACEN) %>%
  create_var_from_codelist(metacore, input_var = ETHNIC, out_var = ETHNICN) %>%
  create_var_from_codelist(metacore, input_var = ARMCD, out_var = COHORT) %>%
  create_var_from_codelist(metacore, input_var = ARMCD, out_var = COHORTC) %>%
  create_var_from_codelist(metacore, input_var = ARM, out_var = ARMN) %>%
  create_var_from_codelist(metacore, input_var = ACTARM, out_var = ACTARMN) %>%
  create_var_from_codelist(metacore, input_var = COUNTRY, out_var = COUNTRYN) %>%
  create_var_from_codelist(metacore, input_var = COUNTRY, out_var = COUNTRYL) %>%
  mutate(
    STUDYIDN = as.numeric(word(USUBJID, 1, sep = fixed("-"))),
    SITEIDN = as.numeric(word(USUBJID, 2, sep = fixed("-"))),
    USUBJIDN = as.numeric(word(USUBJID, 3, sep = fixed("-"))),
    SUBJIDN = as.numeric(SUBJID),
    ROUTE = unique(adex$EXROUTE)[1],
    FORM = unique(adex$EXDOSFRM)[1],
    REGION1 = COUNTRY,
    REGION1N = COUNTRYN,
  ) %>%
  create_var_from_codelist(metacore, input_var = FORM, out_var = FORMN) %>%
  create_var_from_codelist(metacore, input_var = ROUTE, out_var = ROUTEN)
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(covar %>% select(USUBJID, SEX, SEXN, RACE, RACEN, FORM, FORMN))
```

### Derive Additional Baselines

Next we add additional baselines from vital signs and laboratory data.

```{r}
#| label: Baselines
# ---- Derive additional baselines from ADVS and ADLB ----

labsbl <- adlb %>%
  filter(ABLFL == "Y" & PARAMCD %in% c("CREAT", "ALT", "AST", "BILI")) %>%
  mutate(PARAMCDB = paste0(PARAMCD, "BL")) %>%
  select(STUDYID, USUBJID, PARAMCDB, AVAL)

covar_vslb <- covar %>%
  derive_vars_merged(
    dataset_add = advs,
    filter_add = PARAMCD == "HEIGHT" & ABLFL == "Y",
    by_vars = exprs(STUDYID, USUBJID),
    new_vars = exprs(HTBL = AVAL)
  ) %>%
  derive_vars_merged(
    dataset_add = advs,
    filter_add = PARAMCD == "WEIGHT" & ABLFL == "Y",
    by_vars = exprs(STUDYID, USUBJID),
    new_vars = exprs(WTBL = AVAL)
  ) %>%
  derive_vars_transposed(
    dataset_merge = labsbl,
    by_vars = exprs(STUDYID, USUBJID),
    key_var = PARAMCDB,
    value_var = AVAL
  ) %>%
  mutate(
    BMIBL = compute_bmi(height = HTBL, weight = WTBL),
    BSABL = compute_bsa(
      height = HTBL,
      weight = WTBL,
      method = "Mosteller"
    ),
    CRCLBL = compute_egfr(
      creat = CREATBL, creatu = "SI", age = AGE, weight = WTBL, sex = SEX,
      method = "CRCL"
    ),
    EGFRBL = compute_egfr(
      creat = CREATBL, creatu = "SI", age = AGE, weight = WTBL, sex = SEX,
      method = "CKD-EPI"
    )
  ) %>%
  rename(TBILBL = BILIBL)
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(covar_vslb %>% select(USUBJID, WTBL, HTBL, BMIBL, CRCLBL))
```

## Add Exposure Metrics from `ADPP`

An essential component of ER modeling is the exposure metrics calculated from the PK concentration data in `ADPC` or `ADPPK`. Here we will use `AUCLST` and `CMAX` from the `ADPP` template from `{admiral}` in `{pharmaverseadam}`. See the `{aNCA}` package for details about calculating `AUC`, `CMAX`, and other parameters.

```{r}
#| label: AUCSS
# ---- Add Exposure Metrics  ----
covar_auc <- covar_vslb %>%
  derive_vars_transposed(
    dataset_merge = adpp,
    filter = PARAMCD %in% c("AUCLST", "CMAX"),
    by_vars = get_admiral_option("subject_keys"),
    key_var = PARAMCD,
    value_var = AVAL
  ) %>%
  rename(AUCSS = AUCLST, CMAXSS = CMAX)
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(covar_auc %>% select(
  USUBJID, TRT01A, SEX, WTBL, HTBL, AUCSS, CMAXSS
))
```

# Code for Individual Datasets

Derivations specific to each dataset continue below. Select the appropriate tab to view.

## Dataset-Specific Derivations

::: {.panel-tabset}

## ADER {#ader}

### Derive Time to Event Variables from `ADTTE`

We will use parameters from from `ADTTE` for Overall Survival, Progression Free Survival and Duration of Response. We will use `admiral::derive_vars_transposed()` to transpose the `AVAL` for each `PARAMCD`.

```{r}
#| label: ADER Add ADTTE variables
# ---- Derivations ----
# For ADTTE censor variables add "IND" to PARAMCD
adttei <- adtte %>%
  mutate(PARAMCD = paste0(PARAMCD, "IND"))

ader_tte <- adsl %>%
  select(!!!get_admiral_option("subject_keys")) %>%
  # Create OS and PFS variables from ADTTE
  derive_vars_transposed(
    dataset_merge = adtte,
    by_vars = get_admiral_option("subject_keys"),
    key_var = PARAMCD,
    value_var = AVAL
  ) %>%
  # Create OS and PFS censor variables
  derive_vars_transposed(
    dataset_merge = adttei,
    by_vars = get_admiral_option("subject_keys"),
    key_var = PARAMCD,
    value_var = CNSR
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(ader_tte %>% select(USUBJID, OS, OSIND, PFS, PFSIND, RSD, RSDIND))
```

We will use Best Overall Response by Investigator (BOR) from `ADRS`. We attach this to the data with `admiral::derive_vars_merged()`. Note that `derive_vars_transposed()` and `derive_vars_merged()` will be our primary ways of combining data from the different `{admiral}` and `{admiralonco}` templates.

```{r}
#| label: ADER Add ADRS
# ---- Add ADRS data ----
# Add response date to ADSL for duration of response calculation
ader_bor <- ader_tte %>%
  derive_vars_merged(
    dataset_add = adrs,
    filter_add = PARAMCD == "BOR" & ANL01FL == "Y",
    by_vars = get_admiral_option("subject_keys"),
    new_vars = exprs(BOR = AVAL, BORC = AVALC)
  )
```

Add Analysis Sequence Number using `admiral::derive_var_obs_number()`.

```{r}
#| label: ADER ASEQ
# ---- Add Sequence Number ----
ader_aseq <- ader_bor %>%
  derive_var_obs_number(
    by_vars = get_admiral_option("subject_keys"),
    check_type = "error"
  )
```

```{r eval=TRUE, echo=FALSE, purl=FALSE}
print_df(ader_bor %>% select(
  USUBJID, BOR, BORC
))
```

### Combine with Covariates

We combine our covariates with the rest of the data

```{r}
#| label: ADER Combine with Covariates
# Combine covariates with ADER data

ader_prefinal <- ader_aseq %>%
  derive_vars_merged(
    dataset_add = covar_auc,
    by_vars = exprs(STUDYID, USUBJID)
  )
```

### Check Data With metacore and metatools

We use `{metacore}` objects with `{metatools}` functions to perform a number of checks on the data. We will drop variables not in the specs and make sure all the variables from the specs are included.

```{r}
#| label: ADER Metacore
#| warning: false

ader <- ader_prefinal %>%
  drop_unspec_vars(metacore) %>% # Drop unspecified variables from specs
  check_variables(metacore) %>% # Check all variables specified are present and no more
  check_ct_data(metacore) %>% # Checks all variables with CT only contain values within the CT
  order_cols(metacore) %>% # Orders the columns according to the spec
  sort_by_key(metacore) # Sorts the rows by the sort keys
```

### Apply Labels and Formats with xportr

Using {xportr} we check variable type, assign variable length, add variable labels, add variable formats, and save a transport file with `xportr::xportr_write()`.

```{r}
#| label: ADER xportr
dir <- tempdir() # Change to whichever directory you want to save the dataset in

ader_xpt <- ader %>%
  xportr_type(metacore, domain = "ADER") %>% # Coerce variable type to match spec
  xportr_length(metacore) %>% # Assigns SAS length from a variable level metadata
  xportr_label(metacore) %>% # Assigns variable label from metacore specifications
  xportr_format(metacore) %>% # Assigns variable format from metacore specifications
  xportr_df_label(metacore) %>% # Assigns dataset label from metacore specifications
  xportr_write(file.path(dir, "ader.xpt")) # Write xpt v5 transport file
```

## ADEE {#adee}

### Load Specifications
```{r echo=TRUE, message=FALSE}
#| label: ADEE Load Specs
#| warning: false
metacore <- spec_to_metacore("./metadata/pk_spec.xlsx") %>%
  select_dataset("ADEE")
```

```{r}
#| label: ADEE Base dataset
# ---- Create adee base dataset

# Get variable names from both datasets
adsl_vars <- names(adsl)
adtte_vars <- names(adtte)

# Find common variables
common_vars <- intersect(adsl_vars, adtte_vars)

# Remove key variables to get variables to drop
vars_to_drop <- setdiff(common_vars, c("STUDYID", "USUBJID"))

# Ensure PARAMN exists in ADTTE
if (!"PARAMN" %in% names(adtte)) {
  adtte <- adtte %>%
    mutate(
      PARAMN = case_when(
        PARAMCD == "PFS" ~ 1,
        PARAMCD == "OS" ~ 2,
        PARAMCD == "TTP" ~ 3,
        PARAMCD == "TTNT" ~ 4,
        TRUE ~ 99
      )
    )
}

# ---- Create ADEE Base

adee_base <- adtte %>%
  # Filter to efficacy endpoints
  filter(PARAMCD %in% c("OS", "PFS", "TTP", "TTNT")) %>%
  # Add derived variables
  mutate(
    EVENT = 1 - CNSR,
    AVALU = if_else(!is.na(AVAL), "DAYS", NA_character_),
  ) %>%
  # Remove overlapping variables (use clean method)
  select(-any_of(vars_to_drop))
```

### Add Analysis Variables

```{r}
#| label: ADEE Analysis variables
# ---- Add Analysis variables

adee_aseq <- adee_base %>%
  # Analysis flags
  mutate(
    ANL01FL = if_else(PARAMCD == "PFS", "Y", ""),
    ANL02FL = if_else(PARAMCD == "OS", "Y", ""),
    ANL03FL = if_else(PARAMCD == "TTP", "Y", ""),
    ANL04FL = if_else(PARAMCD == "TTNT", "Y", "")
  ) %>%
  # Parameter categories
  mutate(
    PARCAT1 = "EFFICACY",
    PARCAT2 = "TIME TO EVENT"
  ) %>%
  # Sequence number
  derive_var_obs_number(
    by_vars = exprs(STUDYID, USUBJID),
    order = exprs(PARAMCD),
    new_var = ASEQ,
    check_type = "error"
  )
```

### Combine with Covariates and Exposure

We combine our covariates with the rest of the data

```{r}
#| label: ADEE Combine with Covariates
# Combine covariates with ADER data

adee_prefinal <- adee_aseq %>%
  derive_vars_merged(
    dataset_add = covar_auc,
    by_vars = exprs(STUDYID, USUBJID)
  )
```



### Check Data With metacore and metatools

We use `{metacore}` objects with `{metatools}` functions to perform a number of checks on the data. We will drop variables not in the specs and make sure all the variables from the specs are included.

```{r}
#| label: Metacore
#| warning: false

adee <- adee_prefinal %>%
  drop_unspec_vars(metacore) %>% # Drop unspecified variables from specs
  check_variables(metacore) %>% # Check all variables specified are present and no more
  check_ct_data(metacore) %>% # Checks all variables with CT only contain values within the CT
  order_cols(metacore) %>% # Orders the columns according to the spec
  sort_by_key(metacore) # Sorts the rows by the sort keys
```

### Apply Labels and Formats with xportr

Using {xportr} we check variable type, assign variable length, add variable labels, add variable formats, and save a transport file with `xportr::xportr_write()`.

```{r}
#| label: xportr
dir <- tempdir() # Change to whichever directory you want to save the dataset in

adee_xpt <- adee %>%
  xportr_type(metacore, domain = "ADEE") %>% # Coerce variable type to match spec
  xportr_length(metacore) %>% # Assigns SAS length from a variable level metadata
  xportr_label(metacore) %>% # Assigns variable label from metacore specifications
  xportr_format(metacore) %>% # Assigns variable format from metacore specifications
  xportr_df_label(metacore) %>% # Assigns dataset label from metacore specifications
  xportr_write(file.path(dir, "adee.xpt")) # Write xpt v5 transport file
```


## ADES {#ades}

```{r}
# [Similar structure...]
```

## ADTRR {#adtrr}

```{r}
# [Similar structure...]
```


:::
