---
title: "ADRS"
order: 4
---

```{r setup script, include=FALSE, purl=FALSE}
knitr::knit_hooks$set(purl = knitr::hook_purl)
```

# Introduction

The Response Analysis Dataset (`ADRS`) is an essential part of oncology clinical trials for monitoring disease response and progression. This article describes how to create an `ADRS` ADaM dataset, focusing on common oncology endpoint parameters based on RECIST v1.1 criteria. The primary response values include `CR` (Complete Response), `PR` (Partial Response), `SD` (Stable Disease), `NON-CR/NON-PD` (Non-CR/Non-PD), `PD` (Progressive Disease), and `NE` (Not Evaluable).

This guide uses key pharmaverse packages along with tidyverse components to demonstrate the step-by-step process, ensuring the inclusion of metadata, validation, and exporting to a compliant SAS transport file (XPT).

# Load Packages

First, we load the necessary packages required for creating the `ADRS` dataset.

```{r setup, message=FALSE, warning=FALSE, results='hold'}
library(admiral)
library(admiralonco)
library(metacore)
library(metatools)
library(xportr)
library(pharmaversesdtm)
library(pharmaverseadam)
library(dplyr)
library(lubridate)
library(stringr)
```

# Load Specifications for Metacore

Load the specifications stored in an Excel file into a `{metacore}` object.

```{r read-specs}
#| warning: false
metacore <- spec_to_metacore("../metadata/onco_spec.xlsx") %>%
  select_dataset("ADRS")
```

# Load Source Datasets

We load the required SDTM datasets (`RS`, `TU`) and ADaM dataset (`ADSL`) necessary for creating the `ADRS` dataset.

```{r load-data}
data("adsl")
data("rs_onco_recist")
data("tu_onco_recist")

rs <- rs_onco_recist
tu <- tu_onco_recist

# Convert blanks to NA
rs <- convert_blanks_to_na(rs)
tu <- convert_blanks_to_na(tu)
```

# Merging ADSL with RS

Merge `ADSL` to the `RS` domain by selecting only the necessary variables for derivations.

```{r merge-adsl-rs}
adsl_vars <- exprs(RANDDT)
adrs_merged <- derive_vars_merged(
  rs,
  dataset_add = adsl,
  new_vars = adsl_vars,
  by_vars = exprs(STUDYID, USUBJID)
)
head(adrs_merged, n=10)
```

# Pre-processing of Input Records

## Select Overall Response Records and Set Parameter Details

Filter the `RS` domain to include only overall response records assessed by the investigator and set the parameter details accordingly.

```{r set-param-details}
adrs_ovr <- adrs_merged %>%
  filter(RSEVAL == "INVESTIGATOR" & RSTESTCD == "OVRLRESP") %>%
  mutate(
    PARAMCD = "OVR",
    PARAM = "Overall Response by Investigator",
    PARCAT1 = "Tumor Response",
    PARCAT2 = "Investigator",
    PARCAT3 = "RECIST 1.1"
  )
head(adrs_ovr, n=10)
```

## Date Imputation and Deriving `ADT`, `ADTF`, `AVISIT`

Impute missing dates and derive analysis dates and visits.

```{r impute-dates}
adrs_imputed <- adrs_ovr %>%
  derive_vars_dt(
    dtc = RSDTC,
    new_vars_prefix = "A",
    highest_imputation = "D",
    date_imputation = "last"
  ) %>%
  mutate(AVISIT = VISIT)
head(adrs_imputed, n=10)
```

## Derive `AVALC` and `AVAL`

Populate `AVALC` with assessed values and create the numeric version `AVAL`.

```{r derive-aval}
adrs_aval <- adrs_imputed %>%
  mutate(
    AVALC = RSSTRESC,
    AVAL = aval_resp(AVALC)
  )
head(adrs_aval, n=10)
```

## Flag Worst Assessment at Each Date (`ANL01FL`)

Flag the worst assessment at each date, considering only valid assessments from the randomization date onward.

```{r flag-worst-assessment}
worst_resp <- function(arg) {
  case_when(
    arg == "NE" ~ 1,
    arg == "CR" ~ 2,
    arg == "PR" ~ 3,
    arg == "SD" ~ 4,
    arg == "NON-CR/NON-PD" ~ 5,
    arg == "PD" ~ 6,
    TRUE ~ 0
  )
}

adrs_anl01fl <- adrs_aval %>%
  restrict_derivation(
    derivation = derive_var_extreme_flag,
    args = params(
      by_vars = exprs(STUDYID, USUBJID, ADT),
      order = exprs(worst_resp(AVALC), RSSEQ),
      new_var = ANL01FL,
      mode = "last"
    ),
    filter = !is.na(AVAL) & ADT >= RANDDT
  )
head(adrs_anl01fl, n=10)
```

# Deriving Parameters

## Derive Progressive Disease Parameter

Use the `derive_extreme_records()` function to find the date of first `PD`.

```{r derive-pd}
adrs_pd <- adrs_anl01fl %>%
  derive_extreme_records(
    dataset_ref = adsl,
    dataset_add = adrs_anl01fl,
    by_vars = exprs(STUDYID, USUBJID),
    filter_add = PARAMCD == "OVR" & AVALC == "PD" & ANL01FL == "Y",
    order = exprs(ADT, RSSEQ),
    mode = "first",
    exist_flag = AVALC,
    false_value = "N",
    set_values_to = exprs(
      PARAMCD = "PD",
      PARAM = "Disease Progression by Investigator",
      PARCAT1 = "Tumor Response",
      PARCAT2 = "Investigator",
      PARCAT3 = "RECIST 1.1",
      AVAL = yn_to_numeric(AVALC),
      ANL01FL = "Y"
    )
  )
head(adrs_pd, n=10)
```

## Derive Death Parameter

Create a new death parameter using the death date from `ADSL`.

```{r derive-death}
adsldth <- adsl %>%
  select(STUDYID, USUBJID, DTHDT, !!!adsl_vars)

adrs_death <- adrs_pd %>%
  derive_extreme_records(
    dataset_ref = adsldth,
    dataset_add = adsldth,
    by_vars = exprs(STUDYID, USUBJID),
    filter_add = !is.na(DTHDT),
    exist_flag = AVALC,
    false_value = "N",
    set_values_to = exprs(
      PARAMCD = "DEATH",
      PARAM = "Death",
      PARCAT1 = "Reference Event",
      AVAL = yn_to_numeric(AVALC),
      ANL01FL = "Y",
      ADT = DTHDT
    )
  ) %>%
  select(-DTHDT)
head(adrs_death, n=10)
```

## Derive Last Disease Assessment Parameter

Create a parameter for the last disease assessment.

```{r derive-lsta}
adrs_lsta <- adrs_death %>%
  derive_extreme_records(
    dataset_ref = adsl,
    dataset_add = adrs_death,
    by_vars = exprs(STUDYID, USUBJID),
    filter_add = PARAMCD == "OVR" & ANL01FL == "Y",
    order = exprs(ADT, RSSEQ),
    mode = "last",
    set_values_to = exprs(
      PARAMCD = "LSTA",
      PARAM = "Last Disease Assessment by Investigator",
      PARCAT1 = "Tumor Response",
      PARCAT2 = "Investigator",
      PARCAT3 = "RECIST 1.1",
      ANL01FL = "Y"
    )
  )
head(adrs_lsta, n=10)
```

# Apply Metadata and Perform Associated Checks

Apply metadata and conduct checks to ensure data quality and compliance.

```{r apply-metadata-check, eval=FALSE}
adrs_checked <- adrs_lsta %>%
  drop_unspec_vars(metacore) %>% # Drop variables not specified in metadata
  check_variables(metacore) %>% # Check all variables specified are present and no more
  check_ct_data(metacore) %>% # Check controlled terminology
  order_cols(metacore) %>% # Order columns according to metadata
  sort_by_key(metacore) # Sort rows by sort keys
```

# Apply Labels and Formats with xportr

Finally, apply labels, formats, and export the dataset to an XPT file.

```{r xportr}
dir <- tempdir() # Specify the directory for saving the XPT file

adrs_xpt <- adrs_lsta %>%
  xportr_type(metacore, domain = "ADRS") %>% # Coerce variable types to match metadata
  xportr_length(metacore) %>% # Assign variable lengths from metadata
  xportr_label(metacore) %>% # Assign variable labels from metadata
  xportr_format(metacore) %>% # Assign variable formats from metadata
  xportr_df_label(metacore) %>% # Assign dataset labels from metadata
  xportr_write(file.path(dir, "adrs.xpt")) # Write the XPT file
```
