---
title: "ADSL"
---

## Introduction

This guide will show you how four pharmaverse packages, along with some from
tidyverse, can be used to create an ADaM such as `ADSL` end-to-end, using
`{pharmaversesdtm}` SDTM data as input.

The four packages used with a brief description of their purpose are as follows:

* [`{metacore}`](https://atorus-research.github.io/metacore/): provides harmonized
metadata/specifications object.
* [`{metatools}`](https://pharmaverse.github.io/metatools/): uses the provided
metadata to build/enhance and check the dataset.
* [`{admiral}`](https://pharmaverse.github.io/admiral/index.html): provides the
ADaM derivations.
* [`{xportr}`](https://atorus-research.github.io/xportr/): delivers the SAS
transport file (XPT) and eSub checks.

It is important to understand `{metacore}` objects by reading through the above
linked package site, as these are fundamental to being able to use `{metatools}`
and `{xportr}`. Each company may need to build a specification reader to create
these objects from their source standard specification templates.

## Load Data and Required pharmaverse Packages

The first step is to load our pharmaverse packages and input data.

```{r setup, message=FALSE, warning=FALSE, results='hold'}
library(metacore)
library(metatools)
library(pharmaversesdtm)
library(admiral)
library(xportr)
library(dplyr)
library(tidyr)
library(lubridate)
library(stringr)

# Read in input SDTM data 
data("dm")
data("ex")
```

Next we need to load the specification file in the form of a `{metacore}` object.

```{r metacore, warning=FALSE, results='hold'}
# Read in metacore object 
load(metacore_example("pilot_ADaM.rda"))
metacore <- metacore %>% 
   select_dataset("ADSL")
```

Here is an example of how a `{metacore}` object looks showing variable level
metadata:

```{r}
metacore$ds_vars
```
  
## Start Building Derivations

The first derivation step we are going to do is to pull through all the columns
that come directly from the SDTM datasets. You might know which datasets you are
going to pull from directly already, but if you don't you can call
`metatools::build_from_derived()` with just an empty list and the error will tell
you which datasets you need to supply. 

```{r, error=TRUE}
build_from_derived(metacore, list(), predecessor_only = FALSE)
```

In this case all the columns come from `DM` so that is the only dataset we will
pass into `metatools::build_from_derived()`. The resulting dataset has all the
columns combined and any columns that needed renaming between SDTM and ADaM are
renamed.

```{r demographcis}
adsl_preds <- build_from_derived(metacore, 
                                 ds_list = list("dm" = dm), 
                                 predecessor_only = FALSE, keep = TRUE)
head(adsl_preds, n=10)
```

Now we have the base dataset, we can start to create some variables. We can start
with creating the subgroups using the controlled terminology, in this case `AGEGR1`.
The metacore object holds all the metadata needed to make `ADSL`. Part of that
metadata is the controlled terminology, which can help automate the creation of
subgroups. We can look into the `{metacore}` object and see the controlled
terminology for `AGEGR1`.

```{r}
get_control_term(metacore, variable = AGEGR1)
```

Because this controlled terminology is written in a fairly standard format we
can automate the creation of `AGEGR1`. The function `metatools::create_cat_var()`
takes in a `{metacore}` object, a reference variable - in this case `AGE` because
that is the continuous variable `AGEGR1` is created from, and the name of the
sub-grouped variable. It will take the controlled terminology from the sub-grouped
variable and group the reference variables accordingly. 

Using a similar philosophy we can create the numeric version of `RACE` using the
controlled terminology stored in the `{metacore}` object with the
`metatools::create_var_from_codelist()` function.

```{r ct}
adsl_ct <- adsl_preds %>% 
   create_cat_var(metacore, ref_var = AGE, 
                  grp_var = AGEGR1, num_grp_var = AGEGR1N) %>% 
   create_var_from_codelist(metacore = metacore, 
                            input_var = RACE, 
                            out_var = RACEN) %>% 
   #Removing screen failures from ARM and TRT01P to match the define and FDA guidence
   mutate(ARM = if_else(ARM == "Screen Failure", NA_character_, ARM),
          TRT01P = if_else(TRT01P == "Screen Failure", NA_character_, TRT01P)
   )

head(adsl_ct, n=10)
```

Now we have sorted out what we can easily do with controlled terminology it is
time to start deriving some variables.
Here you could refer directly to using the `{admiral}` template and [vignette](https://pharmaverse.github.io/admiral/cran-release/articles/adsl.html)
in practice, but for the purpose of this end-to-end ADaM vignette we will share
a few exposure derivations from there.
We derive the start and end of treatment (which requires dates to first be
converted from DTC to DTM), the treatment duration, and the safety population flag.

```{r exposure}
ex_ext <- ex %>%
  derive_vars_dtm(
    dtc = EXSTDTC,
    new_vars_prefix = "EXST"
  ) %>%
  derive_vars_dtm(
    dtc = EXENDTC,
    new_vars_prefix = "EXEN",
    time_imputation = "last"
  )

adsl_raw <- adsl_ct %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 |
      (EXDOSE == 0 &
        str_detect(EXTRT, "PLACEBO"))) & nchar(EXSTDTC) >= 10,
    new_vars = exprs(TRTSDTM = EXSTDTM),
    order = exprs(EXSTDTM, EXSEQ),
    mode = "first",
    by_vars = exprs(STUDYID, USUBJID)
  ) %>%
  derive_vars_merged(
    dataset_add = ex_ext,
    filter_add = (EXDOSE > 0 |
      (EXDOSE == 0 &
        str_detect(EXTRT, "PLACEBO"))) & nchar(EXENDTC) >= 10,
    new_vars = exprs(TRTEDTM = EXENDTM),
    order = exprs(EXENDTM, EXSEQ),
    mode = "last",
    by_vars = exprs(STUDYID, USUBJID)
  ) %>%
   derive_vars_dtm_to_dt(source_vars = exprs(TRTSDTM, TRTEDTM)) %>%  #Convert Datetime variables to date 
   derive_var_trtdurd() %>% 
   derive_var_merged_exist_flag(
     dataset_add = ex,
     by_vars = exprs(STUDYID, USUBJID),
     new_var = SAFFL,
     condition = (EXDOSE > 0 | (EXDOSE == 0 & str_detect(EXTRT, "PLACEBO")))
   ) %>% 
   drop_unspec_vars(metacore) #This will drop any columns that aren't specified in the metacore object

head(adsl_raw, n=10)
```

```{r, warning=FALSE, message=FALSE, include=FALSE}
# Create dummy variables to match metacore specs to avoid later errors
# In practice these would be mainly created using derivation functions from admiral
adsl_raw <- adsl_raw %>%
  mutate(
    SITEGR1 = NA,
    TRT01PN = NA,
    TRT01A = NA,
    TRT01AN = NA,
    AVGDD = NA,
    CUMDOSE = NA,
    ITTFL = NA,
    EFFFL = NA,
    COMP8FL = NA,
    COMP16FL = NA,
    COMP24FL = NA,
    DISCONFL = NA,
    DSRAEFL = NA,
    BMIBL = NA,
    BMIBLGR1 = NA,
    HEIGHTBL = NA,
    WEIGHTBL = NA,
    EDUCLVL = NA,
    DISONSDT = NA,
    DURDIS = NA,
    DURDSGR1 = NA,
    VISIT1DT = NA,
    VISNUMEN = NA,
    RFENDT = NA,
    DCDECOD = NA,
    EOSSTT = NA,
    DCSREAS = NA,
    MMSETOT = NA
  )
```

## Apply Metadata to Create an eSub XPT and Perform Associated Checks

Now we have all the variables defined we can run some checks before applying the
necessary formatting.
The top four functions performing checks and sorting/ordering come from
`{metatools}`, whereas the others focused around applying attributes to prepare
for XPT come from `{xportr}`. At the end you could add a call to
`xportr::xportr_write()` to produce the XPT file.

```{r checks, warning=FALSE, message=FALSE}

adsl_raw %>% 
   check_variables(metacore) %>% # Check all variables specified are present and no more
   check_ct_data(metacore, na_acceptable = TRUE) %>% # Checks all variables with CT only contain values within the CT
   order_cols(metacore) %>% # Orders the columns according to the spec
   sort_by_key(metacore) %>% # Sorts the rows by the sort keys 
   xportr_type(metacore) %>% # Coerce variable type to match spec
   xportr_length(metacore) %>% # Assigns SAS length from a variable level metadata 
   xportr_label(metacore) %>% # Assigns variable label from metacore specifications 
   xportr_df_label(metacore) # Assigns dataset label from metacore specifications
```
